---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "Number of accidents per day"
author:
    - "Valeria Iapaolo"
    - "Oswaldo Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: true
    pdf_document:
        toc: true
        toc_depth: 3
        number_section: true
date: "2023-11-10"
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center',
    #fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r libraries inclusions, include=FALSE}
library(tidyverse)
library(mgcv)
library(splines)
library(lubridate)
library(ggplot2)
library(conformalInference)
library(knitr)
library(robustbase)
```
Loading the dataset
```{r}
load("C:/Users/ipval/Desktop/Dataset NP/full_collisions.RData")

```
Let's have a look
```{r}
glimpse(full_collisions)
```
The aim of this section is to analyse the number of collisions per day. So we begin by computing this quantity.

## Local authority district
Let's look first to the local authority district
```{r}

full_collisions <- full_collisions[!is.na(full_collisions$local_authority_district), ]


full_collisions %>% ggplot(aes(local_authority_district)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
nlevels(full_collisions$local_authority_district)
```
There are 416 districts.

## Wheather conditions
Now we have a look to the weather conditions
```{r}
levels(full_collisions$weather_conditions)
full_collisions <- full_collisions %>%
  filter(!(weather_conditions %in% c("Data missing or out of range", "Unknown","Other")))
table(full_collisions$weather_conditions)
```
Now we create 2 new variables called 'wind' and 'rainfall' that specify the weather conditions.
```{r}

full_collisions <- full_collisions %>%
  mutate(
    wind = case_when(
      weather_conditions %in% c("Fine + high winds", "Raining + high winds",
                                             "Snowing + high winds") ~ "High winds",
      weather_conditions %in% c("Fine no high winds", "Raining no high winds",
                                             "Snowing no high winds","Fog or mist") ~ "No high winds"
      
    ),
    rainfall = case_when(
      weather_conditions %in% c("Raining + high winds", "Raining no high winds") ~ "Raining",
      weather_conditions %in% c("Snowing + high winds", "Snowing no high winds") ~ "Snowing",
      weather_conditions %in% c("Fine + high winds", "Fine no high winds") ~ "Fine",
      weather_conditions %in% c("Fog or mist") ~ "Fog"
    )
  )

```

Now we count the number of accidents per day and per district.
```{r}
number_of_collisions_per_day <- full_collisions %>%
  group_by(datetime_date = as.Date(datetime), local_authority_district) %>%
  summarise(number_of_collisions = n(),
            most_common_wind = names(sort(table(wind), decreasing = TRUE)[1]),
            most_common_rainfall = names(sort(table(rainfall), decreasing = TRUE)[1])
  )


number_of_collisions_per_day <- number_of_collisions_per_day %>%
  mutate(
    year = lubridate::year(datetime_date),
    day_of_the_year = lubridate::yday(datetime_date),
    day_type = ifelse(weekdays(datetime_date) %in% c("sabato", "domenica"), "weekend", "weekday")
  )

number_of_collisions_per_day$day_type<-factor(number_of_collisions_per_day$day_type)
number_of_collisions_per_day$most_common_rainfall<-factor(number_of_collisions_per_day$most_common_rainfall)
number_of_collisions_per_day$most_common_wind<-factor(number_of_collisions_per_day$most_common_wind)

glimpse(number_of_collisions_per_day)
```

Plot of the number of collisions per day and per district

```{r}
specific_data <- number_of_collisions_per_day %>%
   filter(year==2005, local_authority_district%in%levels(number_of_collisions_per_day$local_authority_district)[185:215])

plot_matrix <- ggplot(specific_data, aes(x = day_of_the_year, y = local_authority_district, fill = number_of_collisions)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "blue") +
  labs(x = "Day of the year", y = "Local authority district", fill = "Number of collisions") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  

print(plot_matrix)

```

I distretti in cui avvengono più incidenti sono:
```{r}
levels(number_of_collisions_per_day$local_authority_district)[25]
levels(number_of_collisions_per_day$local_authority_district)[36]
levels(number_of_collisions_per_day$local_authority_district)[145]
levels(number_of_collisions_per_day$local_authority_district)[193]
levels(number_of_collisions_per_day$local_authority_district)[199]
levels(number_of_collisions_per_day$local_authority_district)[206]
levels(number_of_collisions_per_day$local_authority_district)[300]
levels(number_of_collisions_per_day$local_authority_district)[399]
levels(number_of_collisions_per_day$local_authority_district)[405]
```


```{r}
#save(number_of_collisions_per_day,file = "number_of_collisions_per_day.RData")
```

```{r}
load("C:/Users/ipval/Desktop/Dataset NP/number_of_collisions_per_day.RData")
nlevels(number_of_collisions_per_day$local_authority_district)
```
The model is too heavy so we reduce a bit our dataset removing some districts.
```{r}
number_of_collisions_per_day <- number_of_collisions_per_day %>%
   filter(local_authority_district%in%levels(number_of_collisions_per_day$local_authority_district)[1:200],year<2020)
```

Finally, we fit our model

```{r}

#model_gam <- gam(number_of_collisions ~ day_type + most_common_wind + most_common_rainfall +
                # year + s(day_of_the_year, bs='cr') + s(local_authority_district, bs = "re"), 
                # data = number_of_collisions_per_day, family=poisson(), method='REML')


#summary(model_gam)

```

```{r}
#saveRDS(model_gam,file='summary_gam')
```

```{r}
gam_model <- readRDS('summary_gam')
summary(gam_model)
```
According to this model the mean number of accidents in Birmingham on 1/02/2019 is:

```{r}

pred = predict(gam_model,
                     newdata = data.frame(local_authority_district=rep('Birmingham',3),most_common_rainfall=c('Fine','Raining'
                     ,'Snowing'),most_common_wind=rep('No high winds',3),day_type=rep('weekday',3),
                     day_of_the_year=rep(32,3),year=rep(2019,3)))
exp(pred)

Mean_number_of_collisions <- exp(pred)
Rainfall=c('$Fine$','$Rain$','$Snow$')


kable(
  data.frame(Rainfall, Mean_number_of_collisions),
  digits = c(0, 2),
  escape = FALSE,
  align='c'
)

```

```{r}
plot(gam_model)
```
Normality test for the residuals
```{r}
hist(gam_model$residuals)


```

```{r}
plot(gam_model$fitted.values,gam_model$residuals)
```

```{r}
fitted_values <- predict(gam_model, type = "response")  
residuals <- resid(gam_model, type = "response")        

plot(fitted_values, residuals, xlab = "Fitted Values", ylab = "Residuals", main = "Fitted vs Residuals")

```

```{r}
shapiro.test(gam_model$residuals[sample(5000)])
```
```{r}
mu.res <- mean(gam_model$residuals)
s.res <- sd(gam_model$residuals)
x <- as.matrix(gam_model$residuals)
y <- as.matrix( rnorm(length(gam_model$residuals), mu.res, sqrt(s.res)) ) 
DepthProc::ddPlot(x, y,
                  depth_params = list(method='Tukey'))
```



Model assumptions are not satisfied
```{r}
gam.vcomp(gam_model)
```


```{r}
random_effects <- gam_model$coefficients[17:216]
data <- data.frame(category = levels(number_of_collisions_per_day$local_authority_district)[1:200], effect = random_effects)

dotplot <- ggplot(data, aes(x = category, y = effect)) +
  geom_point(size = 1) +
  labs(title = "Dotplot of the Random Effects",
       x = "Categoria",
       y = "Random Effect") + coord_flip()

print(dotplot)

```
```{r}
levels(number_of_collisions_per_day$local_authority_district)[which(random_effects < -0.4)]
levels(number_of_collisions_per_day$local_authority_district)[which.min(random_effects)]
```
Now we use the function "lmrob" to fit the model in a robust way.
We fit the same model as before but without the random intercept for the local authority district.

```{r}

model_robust <- lmrob(number_of_collisions ~ day_type + most_common_wind + most_common_rainfall +
                      year + bs(day_of_the_year),
                      data = number_of_collisions_per_day,
                      init='M-S',
                      control = lmrob.control(fast.s.large.n = Inf))

```

```{r}
summary(model_robust)
```
```{r}
#saveRDS(model_robust,file='model_robust')
```

```{r}
model_robust <- readRDS('model_robust')
summary(model_robust)
```
```{r}
plot(model_robust)
```
```{r}
coefficients(model_robust)
```
```{r}
coefficients(gam_model)
```

