---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "dayly accidents using location and time and random effects"
author:
    - "Valeria Iapaolo"
    - "Oswaldo Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: true
    pdf_document:
        toc: true
        toc_depth: 3
        number_section: true
date: "2023-12-15"
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center'
    #fig.path = 'output/',
    #fig.height = 6,
    #fig.width = 12
)
```

```{r libraries inclusions, include=FALSE}
library(tidyverse)
library(mgcv)
library(splines)
library(lubridate)
library(ggplot2)
library(knitr)
library(sp)
library(rgl)
library(splines)
library(pbapply)
library(devtools)
library(visreg)
library(mgcViz)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/full_collisions.RData")
```

```{r}
glimpse(full_collisions)
```

```{r}
df_spatial <- full_collisions %>% 
  select(c(date,latitude,longitude,accident_year,light_conditions,weather_conditions,day_of_week)) %>%
  drop_na()
rm(full_collisions)
```

```{r}
geom_map <- map_data("world",region = "UK") %>% filter(!(subregion %in% c("Northern Ireland")))

geom_map_no_islands <- geom_map %>% filter(group == 15)
```

```{r}
max_lat <- range(geom_map_no_islands$lat)
max_lon <- range(geom_map_no_islands$long)
```

```{r}
geom_map_red <- geom_map %>% filter(lat <= max_lat[2] & lat >= max_lat[1] &
                                       long <= max_lon[2] & long >= max_lon[1])
```

this will allow is to get a better result since avoiding the extremes:

```{r}
df_spatial <- df_spatial %>% filter(latitude <= max_lat[2] & latitude >= max_lat[1] &
                                       longitude <= max_lon[2] & longitude >= max_lon[1])
```

```{r}
df_binned <- df_spatial %>% mutate(latitude = ceiling(10*latitude)/10,longitude = ceiling(10*longitude)/10) %>% group_by(latitude,longitude,date) %>% summarise(n = n()) %>% ungroup()
```

selecting a reduced dataset for faster experimentation 

```{r}
df_red <- df_binned %>% filter(lubridate::year(date) == 2022)
```

```{r}
world_cities <- read.csv("/Users/macbookpro/Documents/Nonparametric Statisics/Project/raw data/worldcities.csv")
```

selecting the biggest cities:

```{r}
uk_cities <- world_cities %>% filter(country == "United Kingdom")

cities <- uk_cities %>% select(c(city,lat,lng,population)) %>% filter(population > 150000) %>% 
  filter(city != "Belfast")
```

we can add more information about the time:

```{r}
df_red_day <- df_red_day %>% mutate(month = lubridate::month(date),day_type = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "weekend", "weekday")) %>% mutate(day_type = as.factor(day_type))
```

we can try to add more information about the weather ad the district:

```{r}
df_weather <- df_spatial %>%
  filter(!(weather_conditions %in% c("Data missing or out of range", "Unknown","Other"))) %>%
  mutate(
    wind = case_when(
      weather_conditions %in% c("Fine + high winds", "Raining + high winds",
                                             "Snowing + high winds") ~ "High winds",
      weather_conditions %in% c("Fine no high winds", "Raining no high winds",
                                             "Snowing no high winds","Fog or mist") ~ "No high winds"
    ),
    rainfall = case_when(
      weather_conditions %in% c("Raining + high winds", "Raining no high winds") ~ "Raining",
      weather_conditions %in% c("Snowing + high winds", "Snowing no high winds") ~ "Snowing",
      weather_conditions %in% c("Fine + high winds", "Fine no high winds") ~ "Fine",
      weather_conditions %in% c("Fog or mist") ~ "Fog"
    )
  ) 

df_weather <- df_weather %>% mutate(wind = as.factor(wind),rainfall = as.factor(rainfall)) %>%
  mutate(latitude = ceiling(10*latitude)/10,longitude = ceiling(10*longitude)/10) %>%
  group_by(latitude,longitude,date) %>% 
  summarise(n = n(),wind = names(sort(table(wind), decreasing = TRUE)[1]),
            precipitation = names(sort(table(rainfall), decreasing = TRUE)[1])) %>% ungroup()
```

```{r}
df_weather$wind <- factor(df_weather$wind)
df_weather$precipitation <- factor(df_weather$precipitation)
```

```{r}
df_weather <- df_weather %>% mutate(day_of_year = yday(date),day_type = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "weekend", "weekday")) %>% mutate(day_type = as.factor(day_type))
```

```{r}
df_weather_red <- df_weather %>% filter(year(date) == 2022)
```

another possibility is to also add a random effect.

we can do it for police force to start.

```{r}
df_district <- full_collisions %>% 
  select(latitude, longitude, police_force,local_authority_district) %>%
  filter(latitude <= max_lat[2] & latitude >= max_lat[1] &
                                       longitude <= max_lon[2] & longitude >= max_lon[1]) %>%
  mutate(latitude = ceiling(10*latitude)/10,longitude = ceiling(10*longitude)/10) %>%
  group_by(latitude,longitude) %>%
  summarise(police_force = names(sort(table(police_force), decreasing = TRUE)[1]),
            district = names(sort(table(local_authority_district), decreasing = TRUE)[1])) %>%
  ungroup()

df_district <- df_district %>% mutate(police_force = factor(police_force),district = factor(district))
```

```{r}
df_weather <- inner_join(df_weather,df_district)
```

```{r}
save(df_weather,file = "df_weather.RData")
```

```{r}
df_weather %>% ggplot(aes(longitude,latitude)) + geom_point(aes(color = police_force))
```

```{r}
df_district %>% ggplot(aes(longitude,latitude)) + geom_point(aes(color = district)) +
  theme(legend.position = "none")
```

from this resolution we are loosing some districts.

we can start by the police force.

```{r}
df_weather_red2 <- df_weather %>% filter(year(date) == 2022)
```

just police raom effects

```{r}
gam_day_police_qp <- bam(n ~ day_type + precipitation + s(day_of_year,k = 53, bs = "cr") + 
                       s(police_force, bs = "re"), 
                data = df_weather_red2, family=quasipoisson(), method='REML')
```

```{r}
summary(gam_day_police_qp)
```

```{r}
b_pol_qp <- getViz(gam_day_police_qp)
```

```{r}
print(plot(b_pol_qp, allTerms = T), pages = 1)
```

```{r}
check(b_pol_qp)
```

the fitted values are not good

if we also use the location

```{r}
gam_day_police_qp2 <- bam(n ~ day_type + precipitation + s(day_of_year,k = 53, bs = "cr") + 
                       s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)) + 
                         s(police_force, bs = "re"), 
                data = df_weather_red2, family=quasipoisson(), method='REML')
```

```{r}
summary(gam_day_police_qp2)
```

```{r}
b_pol_qp2 <- getViz(gam_day_police_qp2)
```

```{r}
print(plot(b_pol_qp2, select = c(1,3:5)), pages = 1)
```

```{r}
plot(sm(b_pol_qp2, 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

```{r}
check(b_pol_qp2)
```

this is a bit better

we are having some effect of the population in the rendom effects.

looking at the district instead:

```{r}
gam_day_district_qp <- bam(n ~ day_type + precipitation + s(day_of_year,k = 53, bs = "cr") + 
                       s(district, bs = "re"), 
                data = df_weather_red2, family=poisson(), method='REML')
```

```{r}
summary(gam_day_district_qp)
```

```{r}
b_dist_qp <- getViz(gam_day_district_qp)
```

```{r}
print(plot(b_dist_qp, allTerms = T), pages = 1)
```

```{r}
check(b_dist_qp)
```

the fitted values are not good

if we also use the location

```{r}
gam_day_district_qp2 <- bam(n ~ day_type + precipitation + s(day_of_year,k = 53, bs = "cr") + 
                       s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)) + 
                         s(district, bs = "re"), 
                data = df_weather_red2, family=quasipoisson(), method='REML')
```

```{r}
summary(gam_day_district_qp2)
```

not much improvement compared to the only districts

```{r}
b_dist_qp2 <- getViz(gam_day_district_qp2)
```

```{r}
print(plot(b_dist_qp2, select = c(1,3:5)), pages = 1)
```

```{r}
plot(sm(b_dist_qp2, 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

this is to be expected, the improvement is not the best

```{r}
check(b_dist_qp2)
```

we can try to have more years.

```{r}
df_weather <-df_weather %>% mutate(year = year(date),day = factor(wday(date)),
                  covid_year = factor((year(date)==2020)))

df_w_red <- df_weather %>% filter(year >= 2018)
```

starting with 5 years to see the results

```{r}
system.time(
gam_day_district_qp_full <- bam(n ~ day + precipitation + wind + year + covid_year  +
                                  s(day_of_year,k = 53, bs = "cr") + 
                       s(district, bs = "re"), 
                data = df_w_red, family=quasipoisson(), method='REML')
)
```

```{r}
summary(gam_day_district_qp_full)
```

```{r}
b_dist_qp_f <- getViz(gam_day_district_qp_full)
```

```{r}
print(plot(b_dist_qp_f, select = c(1,3:7)), pages = 1)
```

```{r}
print(plot(b_dist_qp_f, select = 2), pages = 1)
```

problem with the normality of the random effects

```{r}
check(b_dist_qp_f)
```

not too bed here.

incorporating also the lat and long

```{r}
system.time(
gam_day_district_qp_full2 <- bam(n ~ day + precipitation + wind + year + covid_year  +
                                  s(day_of_year,k = 53, bs = "cr") + 
                                  s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)) +
                       s(district, bs = "re"), 
                data = df_w_red, family=quasipoisson(), method='REML')
)
```

```{r}
summary(gam_day_district_qp_full2)
```

```{r}
saveRDS(gam_day_district_qp_full2,file='gam_day_district_qp_full2')
```

```{r}
gam_day_district_qp_full2 <- readRDS('gam_day_district_qp_full2')
```

```{r}
b_dist_qp_f2 <- getViz(gam_day_district_qp_full2)
```

```{r}
print(plot(b_dist_qp_f2, select = c(1,4:8)), pages = 1)
```

```{r}
print(plot(b_dist_qp_f2, select = 3), pages = 1)
```

problem with the normality of the random effects, not extremely bad

```{r}
plot(sm(b_dist_qp_f2, 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1) + theme(plot.title = element_text(hjust = 0.5),legend.position = "none")
```

```{r}
check(b_dist_qp_f2)
```

this is defiitely the best model so far, the only thig is to group snow ad fog since
they have the same effect.

```{r}
newd <- with(df_w_red, data.frame(latitude = mean(latitude),
                                  longitude = mean(longitude),
                                  day_of_year = 120,
                                  wind = "No high winds",
                                  precipitation = "Fine",
                                  day = 1,
                                  year = 2022,
                                  covid_year = FALSE,
                                district = levels(district)))

p <- predict(gam_day_district_qp_full2, newd, type = "terms", se.fit = TRUE)
re <- p[["fit"]][ , "s(district)"]
se <- p[["se.fit"]][ , "s(district)"]

data <- data.frame(district = levels(df_w_red$district), effect = re,std.err = se)
```

```{r}
data %>% 
ggplot(aes(effect, fct_reorder(district, effect))) +
  geom_vline(xintercept = 0, color = "gray50", lty = 2, linewidth = 1.2) +
  geom_errorbar(aes(
    xmin = effect - 1.96*std.err,
    xmax = effect + 1.96*std.err),width = 0.5, alpha = 0.7) +
  geom_point(size = 1,colour = "blue") + theme_minimal() + 
  labs(y = "District", x = "Random effect") + ggtitle("Random effects for each district") + 
  theme(legend.position = "none",plot.title = element_text(size = 16,hjust = 0.5),
        axis.text.y = element_blank(),axis.ticks.y = element_blank()) 
```

```{r}
tail(data %>% arrange(effect))
```

this are districts of London.

```{r}
head(data %>% arrange(effect))
```

Broxbourne has 15000 inhabitants. 

this does not consider the latitude and logitude components.

we can do the same but for the police stations:

starting with 5 years to see the results

```{r}
system.time(
gam_day_pol_qp_full <- bam(n ~ day + precipitation + wind + year + covid_year  +
                                  s(day_of_year,k = 53, bs = "cr") + 
                       s(police_force, bs = "re"), 
                data = df_w_red, family=quasipoisson(), method='REML')
)
```

```{r}
summary(gam_day_pol_qp_full)
```

```{r}
b_pol_qp_f <- getViz(gam_day_pol_qp_full)
```

```{r}
print(plot(b_pol_qp_f, select = c(1,3:7)), pages = 1)
```

```{r}
print(plot(b_pol_qp_f, select = 2), pages = 1)
```

problem with the normality of the random effects, the outlier is probably london:

```{r}
check(b_pol_qp_f)
```

not too bed here.

```{r}
newd <- with(df_w_red, data.frame(latitude = mean(latitude),
                                  longitude = mean(longitude),
                                  day_of_year = 120,
                                  wind = "No high winds",
                                  precipitation = "Fine",
                                  day = 1,
                                  year = 2022,
                                  covid_year = FALSE,
                                  police_force = levels(police_force)))

p <- predict(gam_day_pol_qp_full, newd, type = "terms", se.fit = TRUE)
re <- p[["fit"]][ , "s(police_force)"]
se <- p[["se.fit"]][ , "s(police_force)"]

data <- data.frame(police_force = levels(df_weather_red2$police_force), effect = re,std.err = se)
```

```{r}
data %>% 
ggplot(aes(effect, fct_reorder(police_force, effect))) +
  geom_vline(xintercept = 0, color = "gray50", lty = 2, linewidth = 1.2) +
  geom_errorbar(aes(
    xmin = effect - 1.96*std.err,
    xmax = effect + 1.96*std.err),width = 0.5, alpha = 0.7) +
  geom_point(size = 1,colour = "blue") + theme_minimal() + 
  labs(y = "District", x = "Random effect") + ggtitle("Random effects for each police force") + 
  theme(legend.position = "none",plot.title = element_text(size = 16,hjust = 0.5)) 
```

this is reasonable

incorporating also the lat and long

```{r}
system.time(
gam_day_pol_qp_full2 <- bam(n ~ day + precipitation + wind + year + covid_year  +
                                  s(day_of_year,k = 53, bs = "cr") + 
                                  s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)) +
                       s(police_force, bs = "re"), 
                data = df_w_red, family=quasipoisson(), method='REML')
)
```

```{r}
summary(gam_day_pol_qp_full2)
```

```{r}
b_pol_qp_f2 <- getViz(gam_day_pol_qp_full2)
```

```{r}
print(plot(b_pol_qp_f2, select = c(1,4:8)), pages = 1)
```

```{r}
print(plot(b_pol_qp_f2, select = 3), pages = 1)
```

normality of the random effects, not bad

```{r}
plot(sm(b_pol_qp_f2, 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

```{r}
check(b_pol_qp_f2)
```

this is defiitely the best model so far, the only thig is to group snow ad fog since
they have the same effect.

```{r}
newd <- with(df_w_red, data.frame(latitude = mean(latitude),
                                  longitude = mean(longitude),
                                  day_of_year = 120,
                                  wind = "No high winds",
                                  precipitation = "Fine",
                                  day = 1,
                                  year = 2022,
                                  covid_year = FALSE,
                                  police_force = levels(police_force)))

p <- predict(gam_day_pol_qp_full2, newd, type = "terms", se.fit = TRUE)
re <- p[["fit"]][ , "s(police_force)"]
se <- p[["se.fit"]][ , "s(police_force)"]

data <- data.frame(police_force = levels(df_weather$police_force), effect = re,std.err = se)
```

```{r}
data %>% 
ggplot(aes(effect, fct_reorder(police_force, effect))) +
  geom_vline(xintercept = 0, color = "gray50", lty = 2, linewidth = 1.2) +
  geom_errorbar(aes(
    xmin = effect - 1.96*std.err,
    xmax = effect + 1.96*std.err),width = 0.5, alpha = 0.7) +
  geom_point(size = 1,colour = "blue") + theme_minimal() + 
  labs(y = "District", x = "Random effect") + ggtitle("Random effects for each police force") + 
  theme(legend.position = "none",plot.title = element_text(size = 16,hjust = 0.5)) 
```

this does not consider the latitude and logitude components.

```{r}
df_w_red2 <- df_w_red %>% mutate(year = year - 2000) %>% mutate(day = fct_collapse(day,
                                           "mon" = 2,
                                           "tue-thur" = c(3,4,5),
                                           "fri" = 6,
                                           "sat" = 7,
                                           "sun"= 1)) %>% mutate(precipitation = fct_collapse(precipitation,
                                           "Fine" = "Fine",
                                           "Raining" = "Raining",
                                           "Fog/snow" = c("Fog","Snowing")))
```

```{r}
system.time(
gam_day_district_qp_full3 <- bam(n ~ day + precipitation + wind + year + covid_year  +
                                  s(day_of_year,k = 53, bs = "cr") + 
                                  s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)) +
                       s(district, bs = "re"), 
                data = df_w_red2, family=quasipoisson(), method='REML')
)
```

```{r}
summary(gam_day_district_qp_full3)
```

```{r}
print(plot(b_dist_qp_f3, select = c(1,4:8)), pages = 1)
```

```{r}
print(plot(b_dist_qp_f3, select = 3), pages = 1)
```

problem with the normality of the random effects, not extremely bad

```{r}
plot(sm(b_dist_qp_f3, 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + 
  xlab("longitude") + ylab("latitude") + ggtitle("spatial component") + coord_fixed(ratio=1) + theme(plot.title = element_text(hjust = 0.5),legend.position = "none")
```

```{r}
check(b_dist_qp_f3)
```

this is defiitely the best model so far, the only thig is to group snow ad fog since
they have the same effect.

```{r}
newd <- with(df_w_red2, data.frame(latitude = mean(latitude),
                                  longitude = mean(longitude),
                                  day_of_year = 120,
                                  wind = "No high winds",
                                  precipitation = "Fine",
                                  day = 1,
                                  year = 2022,
                                  covid_year = FALSE,
                                district = levels(district)))

p <- predict(gam_day_district_qp_full2, newd, type = "terms", se.fit = TRUE)
re <- p[["fit"]][ , "s(district)"]
se <- p[["se.fit"]][ , "s(district)"]

data <- data.frame(district = levels(df_w_red2$district), effect = re,std.err = se)
```

```{r}
data %>% 
ggplot(aes(effect, fct_reorder(district, effect))) +
  geom_vline(xintercept = 0, color = "gray50", lty = 2, linewidth = 1.2) +
  geom_errorbar(aes(
    xmin = effect - 1.96*std.err,
    xmax = effect + 1.96*std.err),width = 0.5, alpha = 0.7) +
  geom_point(size = 1,colour = "blue") + theme_minimal() + 
  labs(y = "District", x = "Random effect") + ggtitle("Random effects for each district") + 
  theme(legend.position = "none",plot.title = element_text(size = 16,hjust = 0.5),
        axis.text.y = element_blank(),axis.ticks.y = element_blank()) 
```

```{r}
tail(data %>% arrange(effect))
```

this are districts of London.

```{r}
head(data %>% arrange(effect))
```

```{r}
df_w_red3 <- df_w_red %>% mutate(year = year - 2000) %>% mutate(precipitation = fct_collapse(precipitation,
                                           "Fine" = "Fine",
                                           "Raining" = "Raining",
                                           "Fog/snow" = c("Fog","Snowing")))
```

```{r}
system.time(
gam_day_district_qp_full4 <- bam(n ~ day + precipitation + wind + year + covid_year  +
                                  s(day_of_year,k = 53, bs = "cr") + 
                                  s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)) +
                       s(district, bs = "re"), 
                data = df_w_red3, family=quasipoisson(), method='REML')
)
```

```{r}
summary(gam_day_district_qp_full4)
```

```{r}
b_dist_qp_f4 <- getViz(gam_day_district_qp_full4)
```

```{r}
print(plot(b_dist_qp_f4, select = c(1,4:8)), pages = 1)
```

```{r}
print(plot(b_dist_qp_f4, select = 3), pages = 1)
```

problem with the normality of the random effects, not extremely bad

```{r}
plot(sm(b_dist_qp_f4, 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + 
  xlab("longitude") + ylab("latitude") + ggtitle("spatial component") + coord_fixed(ratio=1) + theme(plot.title = element_text(hjust = 0.5),legend.position = "none")
```

```{r}
check(b_dist_qp_f4)
```

this is definitely the best model so far, the only thig is to group snow ad fog since
they have the same effect.

```{r}
newd <- with(df_w_red3, data.frame(latitude = mean(latitude),
                                  longitude = mean(longitude),
                                  day_of_year = 120,
                                  wind = "No high winds",
                                  precipitation = "Fine",
                                  day = 1,
                                  year = 2022,
                                  covid_year = FALSE,
                                district = levels(district)))

p <- predict(gam_day_district_qp_full4, newd, type = "terms", se.fit = TRUE)
re <- p[["fit"]][ , "s(district)"]
se <- p[["se.fit"]][ , "s(district)"]

data <- data.frame(district = levels(df_w_red3$district), effect = re,std.err = se)
```

```{r}
data %>% 
ggplot(aes(effect, fct_reorder(district, effect))) +
  geom_vline(xintercept = 0, color = "gray50", lty = 2, linewidth = 1.2) +
  geom_errorbar(aes(
    xmin = effect - 1.96*std.err,
    xmax = effect + 1.96*std.err),width = 0.5, alpha = 0.7) +
  geom_point(size = 1,colour = "blue") + theme_minimal() + 
  labs(y = "District", x = "Random effect") + ggtitle("Random effects for each district") + 
  theme(legend.position = "none",plot.title = element_text(size = 16,hjust = 0.5),
        axis.text.y = element_blank(),axis.ticks.y = element_blank()) 
```

```{r}
tail(data %>% arrange(effect))
```

this are districts of London.

```{r}
head(data %>% arrange(effect))
```
