---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "dayly accidents using location and time"
author:
    - "Valeria Iapaolo"
    - "Oswaldo Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: true
    pdf_document:
        toc: true
        toc_depth: 3
        number_section: true
date: "2023-12-15"
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center'
    #fig.path = 'output/',
    #fig.height = 6,
    #fig.width = 12
)
```

```{r libraries inclusions, include=FALSE}
library(tidyverse)
library(mgcv)
library(splines)
library(lubridate)
library(ggplot2)
library(knitr)
library(sp)
library(rgl)
library(splines)
library(pbapply)
library(devtools)
library(visreg)
library(mgcViz)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/full_collisions.RData")
```

```{r}
glimpse(full_collisions)
```

```{r}
df_spatial <- full_collisions %>% 
  select(c(date,latitude,longitude,accident_year,light_conditions,weather_conditions,day_of_week)) %>%
  drop_na()
rm(full_collisions)
```

```{r}
geom_map <- map_data("world",region = "UK") %>% filter(!(subregion %in% c("Northern Ireland")))

geom_map_no_islands <- geom_map %>% filter(group == 15)
```

```{r}
max_lat <- range(geom_map_no_islands$lat)
max_lon <- range(geom_map_no_islands$long)
```

```{r}
geom_map_red <- geom_map %>% filter(lat <= max_lat[2] & lat >= max_lat[1] &
                                       long <= max_lon[2] & long >= max_lon[1])
```

this will allow is to get a better result since avoiding the extremes:

```{r}
df_spatial <- df_spatial %>% filter(latitude <= max_lat[2] & latitude >= max_lat[1] &
                                       longitude <= max_lon[2] & longitude >= max_lon[1])
```

```{r}
df_binned <- df_spatial %>% mutate(latitude = ceiling(10*latitude)/10,longitude = ceiling(10*longitude)/10) %>% group_by(latitude,longitude,date) %>% summarise(n = n()) %>% ungroup()
```

selecting a reduced dataset for faster experimentation 

```{r}
df_red <- df_binned %>% filter(lubridate::year(date) == 2022)
```

```{r}
world_cities <- read.csv("/Users/macbookpro/Documents/Nonparametric Statisics/Project/raw data/worldcities.csv")
```

selecting the biggest cities:

```{r}
uk_cities <- world_cities %>% filter(country == "United Kingdom")

cities <- uk_cities %>% select(c(city,lat,lng,population)) %>% filter(population > 150000) %>% 
  filter(city != "Belfast")
```

a first element is to deterimie if there is a difference in the year for the locations:

```{r}
df_feb <- df_binned %>% filter(lubridate::year(date) < 2010) %>% filter(lubridate::month(date) == 2)

df_nov <- df_binned %>% filter(lubridate::year(date) < 2010) %>% filter(lubridate::month(date) == 11)
```

fitting the model of the only space in the different times

```{r}
model_feb <- gam(n ~ s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)), 
                data = df_feb, family=poisson(), method='REML')
```

```{r}
summary(model_feb)
```

```{r}
plot(sm(getViz(model_feb), 1)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

```{r}
model_nov <- gam(n ~ s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)), 
                data = df_nov, family=poisson(), method='REML')
```

```{r}
summary(model_nov)
```

```{r}
plot(sm(getViz(model_nov), 1)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

there is not a big difference, no need tor time and space iteractions.

the baseline model is:

```{r}
mt <- gam(n ~ s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)), 
                data = df_red, family=poisson(), method='REML')
```

```{r}
summary(mt)
```

```{r}
plot(mt,asp = 1)
```

```{r}
plot(sm(getViz(mt), 1)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

```{r}
gam.check(mt)
```

starting only with the day of the year

```{r}
df_red_day <- df_red %>% mutate(day_of_year = lubridate::yday(date))
```

```{r}
gam_dayofyear <- gam(n ~ s(day_of_year,k = 12, bs = "cr") + 
                       s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)), 
                data = df_red_day, family=poisson(), method='REML')
```

```{r}
summary(gam_dayofyear)
```

```{r}
plot(gam_dayofyear)
```

```{r}
plot(sm(getViz(gam_dayofyear), 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

```{r}
gam.check(gam_dayofyear)
```

we can add more information about the time:

```{r}
df_red_day <- df_red_day %>% mutate(month = lubridate::month(date),day_type = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "weekend", "weekday")) %>% mutate(day_type = as.factor(day_type))
```

```{r}
gam_day_we <- gam(n ~ day_type + s(day_of_year,k = 53, bs = "cr") + 
                       s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)), 
                data = df_red_day, family=poisson(), method='REML')
```

```{r}
summary(gam_day_we)
```

```{r}
plot(gam_day_we)
```

```{r}
plot(sm(getViz(gam_day_we), 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

```{r}
gam.check(gam_day_we)
```

still the majority is explained by the latitude ad longitude

we can try to add more information about the weather ad the district:

```{r}
df_weather <- df_spatial %>%
  filter(!(weather_conditions %in% c("Data missing or out of range", "Unknown","Other"))) %>%
  mutate(
    wind = case_when(
      weather_conditions %in% c("Fine + high winds", "Raining + high winds",
                                             "Snowing + high winds") ~ "High winds",
      weather_conditions %in% c("Fine no high winds", "Raining no high winds",
                                             "Snowing no high winds","Fog or mist") ~ "No high winds"
    ),
    rainfall = case_when(
      weather_conditions %in% c("Raining + high winds", "Raining no high winds") ~ "Raining",
      weather_conditions %in% c("Snowing + high winds", "Snowing no high winds") ~ "Snowing",
      weather_conditions %in% c("Fine + high winds", "Fine no high winds") ~ "Fine",
      weather_conditions %in% c("Fog or mist") ~ "Fog"
    )
  ) 

df_weather <- df_weather %>% mutate(wind = as.factor(wind),rainfall = as.factor(rainfall)) %>%
  mutate(latitude = ceiling(10*latitude)/10,longitude = ceiling(10*longitude)/10) %>%
  group_by(latitude,longitude,date) %>% 
  summarise(n = n(),wind = names(sort(table(wind), decreasing = TRUE)[1]),
            precipitation = names(sort(table(rainfall), decreasing = TRUE)[1])) %>% ungroup()
```

```{r}
df_weather$wind <- factor(df_weather$wind)
df_weather$precipitation <- factor(df_weather$precipitation)
```

```{r}
df_weather <- df_weather %>% mutate(day_of_year = yday(date),day_type = ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "weekend", "weekday")) %>% mutate(day_type = as.factor(day_type))
```

```{r}
df_weather_red <- df_weather %>% filter(year(date) == 2022)
```

```{r}
gam_day_weather <- gam(n ~ day_type + precipitation + wind + s(day_of_year,k = 53, bs = "cr") + 
                       s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)), 
                data = df_weather_red, family=poisson(), method='REML')
```

```{r}
summary(gam_day_weather)
```

the performance is even worse than before, the precipitations are not really significant here

the reml on the other hand is better

```{r}
plot(gam_day_weather)
```

```{r}
plot(sm(getViz(gam_day_weather), 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

removing the wind since not significant:

```{r}
gam_day_weather <- gam(n ~ day_type + precipitation + s(day_of_year,k = 53, bs = "cr") + 
                       s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)), 
                data = df_weather_red, family=poisson(), method='REML')
```

```{r}
summary(gam_day_weather)
```

```{r}
plot(gam_day_weather)
```

```{r}
plot(sm(getViz(gam_day_weather), 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

```{r}
gam.check(gam_day_weather)
```

```{r}
AIC(mt,gam_dayofyear,gam_day_we,gam_day_weather)
```

there is an improvement in terms of AIC but not i % of deviance expaied.

we can probably do better by simply increasing the spatial resollutio here.

by improving the ammount of spatial basis we should have a better resuklt, 
since it seem that k iss too low

```{r}
gam_day_weather2 <- gam(n ~ day_type + precipitation + s(day_of_year,k = 53, bs = "cr") + 
                       s(longitude,latitude,bs = "gp",k = 200,m = c(1,0.5)), 
                data = df_weather_red, family=poisson(), method='REML')
```

```{r}
summary(gam_day_weather2)
```

this is better

```{r}
b <- getViz(gam_day_weather2)
```

```{r}
print(plot(b, allTerms = T), pages = 1)
```

```{r}
plot(sm(b, 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

```{r}
check(b)
```
still saying it is too low.

```{r}
pl <- plot(b, allTerms = T) + l_points() + l_fitLine(linetype = 3) + l_fitContour() + 
      l_ciLine(colour = 2) + l_ciBar() + l_fitPoints(size = 1, col = 2) + theme_get() + labs(title = NULL)
pl$empty # FALSE: because we added gamLayers
```

```{r}
print(pl,pages = 4)
```


```{r}
AIC(gam_day_weather2)
```

still a improvement but not massive.

changing the different distributions:

tryig to change to quasipoisson:

```{r}
gam_day_weather_qp <- gam(n ~ day_type + precipitation + s(day_of_year,k = 53, bs = "cr") + 
                       s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)), 
                data = df_weather_red, family=quasipoisson(), method='REML')
```

```{r}
summary(gam_day_weather_qp)
```

```{r}
b_qp <- getViz(gam_day_weather_qp)
```

```{r}
print(plot(b_qp, allTerms = T), pages = 1)
```

```{r}
plot(sm(b_qp, 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + 
  xlab("longitude") + ylab("latitude") + ggtitle("Spatial component") + coord_fixed(ratio=1) + theme(legend.position = "none",plot.title = element_text(hjust = 0.5))
```

```{r}
check(b_qp)
```

this seems better as far as fit goes.

```{r}
check1D(b_qp,"day_of_year") + l_dens(type = "cond", alpha = 0.8) + l_points() + l_rug(alpha = 0.2)
```

```{r}
check1D(b_qp,"day_type") + l_points() + l_rug(alpha = 0.2) 
```

```{r}
check1D(b_qp,"precipitation") + l_points() + l_rug(alpha = 0.2) 
```

```{r}
check2D(b_qp, x1 = "longitude", x2 = "latitude") + l_rug() + l_dens(type="joint", alpha=0.6) + l_points()
```

since this is not a true poisson we don't have an exacy aic

trying negative binomial:

```{r}
gam_day_weather_nb <- gam(n ~ day_type + precipitation + s(day_of_year,k = 53, bs = "cr") + 
                       s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)), 
                data = df_weather_red, family=nb(), method='REML')
```

```{r}
summary(gam_day_weather_nb)
```

```{r}
b_nb <- getViz(gam_day_weather_nb)
```

```{r}
print(plot(b_nb, allTerms = T), pages = 1)
```

```{r}
plot(sm(b_nb, 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

```{r}
check(b_nb)
```

```{r}
AIC(gam_day_weather_nb)
```

this is very similar to the normal poisson case.
the quasipoisson is much better in terms of fit.

twedie distribution:

trying negative binomial:

```{r}
gam_day_weather_tw <- gam(n ~ day_type + precipitation + s(day_of_year,k = 53, bs = "cr") + 
                       s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)), 
                data = df_weather_red, family=tw(), method='REML')
```

```{r}
summary(gam_day_weather_tw)
```

```{r}
b_tw <- getViz(gam_day_weather_tw)
```

```{r}
print(plot(b_tw, allTerms = T), pages = 1)
```

```{r}
plot(sm(b_tw, 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + l_points() +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component") + coord_fixed(ratio=1)
```

```{r}
check(b_tw)
```

```{r}
AIC(gam_day_weather_tw)
```

this is the best for AIC
quasipoisson has bigger reml

we could try to better fit the time domain 

```{r}
df_weather_red <- df_weather_red %>% mutate(day = wday(date)) %>% mutate(day = factor(day))
```

```{r}
gam_day_weather_qp2 <- gam(n ~ day + precipitation + s(day_of_year,k = 53, bs = "cr") + 
                       s(longitude,latitude,bs = "gp",k = 100,m = c(1,0.5)), 
                data = df_weather_red, family=quasipoisson(), method='REML')
```

```{r}
summary(gam_day_weather_qp2)
```

```{r}
b_qp2 <- getViz(gam_day_weather_qp2)
```

```{r}
print(plot(b_qp2, allTerms = T), pages = 1)
```

```{r}
plot(sm(b_qp2, 2)) + l_fitRaster() + l_fitContour() +
    geom_polygon(data = geom_map_red,aes(x=long, y = lat,group=group),
                 fill=NA,color="black",lwd = 1,inherit.aes = F) + 
  theme(legend.position = "none",plot.title = element_text(hjust = 0.5)) +
  xlab("Longitude") + ylab("Latitude") + ggtitle("spatial component",) + coord_fixed(ratio=1) 
```

```{r}
check(b_qp2)
```

this is a little better still not much improvement.