---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "functional cp for each police force"
author:
    - "Valeria Iapaolo"
    - "Oswaldo Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
output: html_notebook
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center',
    #fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r}
library(tidyverse)
library(conformalInference.fd)
library(roahd)
library(lubridate)
library(mgcv)
library(mgcViz)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/full_collisions.RData")
```

```{r}
glimpse(full_collisions)
```

we need to create the dataset in a way that they cotain all of the days and a zero if no crashes happened in that day:

```{r}
all_dates <- unique(full_collisions$date)
all_police_forces <- unique(full_collisions$police_force)

full_combinations <- expand.grid(date = all_dates, police_force = all_police_forces)

crashes_per_day_police <- full_combinations %>%
  left_join(full_collisions %>% group_by(date, police_force) %>% summarize(number_of_crashes = n()),
            by = c("date", "police_force")) %>%
  replace_na(list(number_of_crashes = 0))

glimpse(crashes_per_day_police)
```

```{r}
crashes_per_day_police <- crashes_per_day_police %>% mutate(day_of_year = yday(date),
                                                            day = factor(wday(date)),year = year(date))
```

removing the 29 of february for dimensioality problems in the vectors below

```{r}
crashes_per_day_police <- crashes_per_day_police %>% filter(!(month(date)==2 & day(date)==29))
```

```{r}
train_years <- ceiling(seq(2005,2021,by = 1.5))
cal_years <- seq(2006,2021,3)

df_train <- crashes_per_day_police %>% filter(year %in% train_years) 
df_cal <- crashes_per_day_police %>% filter(year %in% cal_years) 
df_test <- crashes_per_day_police %>% filter(year == 2022)
```

starting with a gam and mixed effects

```{r}
gam_train <- bam(number_of_crashes ~ day + year +
                                  s(day_of_year,k = 53, bs = "cr") + 
                       s(police_force, bs = "re"), 
                data = df_train, family=quasipoisson(), method='REML')
```

```{r}
summary(gam_train)
```

```{r}
b <- getViz(gam_train)
```

```{r}
print(plot(b, allTerms = T), pages = 1)
```

1 is sunday here

```{r}
newd <- with(df_train, data.frame(day_of_year = 120,
                                  day = 1,
                                  year = 2022,
                                  police_force = levels(police_force)))

p <- predict(gam_train, newd, type = "terms", se.fit = TRUE)
re <- p[["fit"]][ , "s(police_force)"]
se <- p[["se.fit"]][ , "s(police_force)"]

data <- data.frame(police_force = levels(df_train$police_force), effect = re,std.err = se)
```

```{r}
data %>% 
ggplot(aes(effect, fct_reorder(police_force, effect))) +
  geom_vline(xintercept = 0, color = "gray50", lty = 2, linewidth = 1.2) +
  geom_errorbar(aes(
    xmin = effect - 1.96*std.err,
    xmax = effect + 1.96*std.err),width = 0.5, alpha = 0.7) +
  geom_point(size = 1,colour = "blue") + theme_minimal() + 
  labs(y = "District", x = "Random effect") + ggtitle("Random effects for each police force") + 
  theme(legend.position = "none",plot.title = element_text(size = 16,hjust = 0.5)) 
```

```{r}
check(b)
```

we now need to compute the bands in a functional case

```{r}
res_train <- cbind(df_train,preds = predict(gam_train, df_train, type = "response")) %>%
  arrange(police_force,date) %>% mutate(absdiff = abs(number_of_crashes-preds))

res_cal <- cbind(df_cal,preds = predict(gam_train, df_cal, type = "response")) %>%
  arrange(police_force,date) %>% mutate(absdiff = abs(number_of_crashes-preds))

res_test <- cbind(df_test,preds = predict(gam_train, df_test, type = "response")) %>%
  arrange(police_force,date) %>% mutate(absdiff = abs(number_of_crashes-preds))
```

```{r}
n_pol <- length(levels(crashes_per_day_police$police_force))

m = 12
l = 6
alpha = 0.1
S = matrix(nrow = n_pol,ncol = 365)
k = rep(0,n_pol)
x = 1:365
line_integral = function(x, y) {
  dx = diff(x)
  end = length(y)
  my = (y[1:(end - 1)] + y[2:end]) / 2
  sum(dx *my)
} 

ceiling((m+1)*(1-alpha))
```

we use the largest value for gamma
H1 = I1

computing the modulation functions:

```{r}
for(i in 1:n_pol){
  ad_train <- res_train %>% filter(police_force == levels(police_force)[i]) %>%
  dplyr::select(absdiff)
  a_mat <- matrix(ad_train$absdiff,byrow = F,nrow = 365)
  ncm = apply(a_mat,1,max)
  den = line_integral(x, ncm)
  s = ncm/den
  S[i,] = s
}
```

computing the k:

```{r}
ceiling((l + 1)*(1-alpha))
```

we use the biggest value for the choice of k

```{r}
for(i in 1:n_pol){
  ad_cal <- res_cal %>% filter(police_force == levels(police_force)[i]) %>%
  dplyr::select(absdiff)
  a_mat <- matrix(ad_cal$absdiff,byrow = F,nrow = 365)
  ncs = apply(a_mat/S[i,],1,max)
  k[i] = max(ncs)
}
```

```{r}
i <- 50

df_filtered <- res_test %>% filter(police_force == levels(police_force)[i]) %>%
  arrange(day_of_year)

df_plot <- data.frame(n = df_filtered$number_of_crashes,pred = df_filtered$preds, 
                      ymax = df_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])

```

this is not great:


```{r}
i <- 6

df_filtered <- res_test %>% filter(police_force == levels(police_force)[i]) %>%
  arrange(day_of_year)

df_plot <- data.frame(n = df_filtered$number_of_crashes,pred = df_filtered$preds, 
                      ymax = df_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])

```


```{r}
i <- 30

df_filtered <- res_test %>% filter(police_force == levels(police_force)[i]) %>%
  arrange(day_of_year)

df_plot <- data.frame(n = df_filtered$number_of_crashes,pred = df_filtered$preds, 
                      ymax = df_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])

```

this are too big, it is due to the fact that we are using a mixd effects model, 
so metropolitan police will contaminate all of the data.

we can fix this by making a gam for each police force:

```{r}
gams <- list()

S = matrix(nrow = n_pol,ncol = 365)
k = rep(0,n_pol)

for(i in 1:n_pol){
  
  df_train_mid <- df_train %>% filter(police_force == levels(police_force)[i])
  df_cal_mid <- df_cal %>% filter(police_force == levels(police_force)[i])
  
  gam <- bam(number_of_crashes ~ day + year +
                                  s(day_of_year,k = 53, bs = "cr"), 
                data = df_train_mid, family=quasipoisson(), method='REML')
  
  res_train_mid <- cbind(df_train_mid,preds = predict(gam, df_train_mid, type = "response")) %>%
  arrange(date) %>% mutate(absdiff = abs(number_of_crashes-preds))

  res_cal_mid <- cbind(df_cal_mid,preds = predict(gam, df_cal_mid, type = "response")) %>%
  arrange(date) %>% mutate(absdiff = abs(number_of_crashes-preds))
  
  
  ad_train <- res_train_mid %>% filter(police_force == levels(police_force)[i]) %>%
  dplyr::select(absdiff)
  a_mat <- matrix(ad_train$absdiff,byrow = F,nrow = 365)
  ncm = apply(a_mat,1,max)
  den = line_integral(x, ncm)
  s = ncm/den
  S[i,] = s
  ad_cal <- res_cal_mid %>% filter(police_force == levels(police_force)[i]) %>%
  dplyr::select(absdiff)
  a_mat <- matrix(ad_cal$absdiff,byrow = F,nrow = 365)
  ncs = apply(a_mat/S[i,],1,max)
  k[i] = max(ncs)
  
  gams[[i]] <- gam
}
```

midlands

```{r}
i <- 50
df_test_filtered <- df_test %>% filter(police_force == levels(police_force)[i])
preds <- predict(gams[[i]], df_test_filtered, type = "response")
df_test_filtered  <- cbind(df_test_filtered,preds) %>% arrange(day_of_year)
```

```{r}
df_plot <- data.frame(n = df_test_filtered$number_of_crashes,pred = df_test_filtered$preds, 
                      ymax = df_test_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_test_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])
```

London:

```{r}
i <- 30
df_test_filtered <- df_test %>% filter(police_force == levels(police_force)[i])
preds <- predict(gams[[i]], df_test_filtered, type = "response")
df_test_filtered  <- cbind(df_test_filtered,preds) %>% arrange(day_of_year)
```

```{r}
df_plot <- data.frame(n = df_test_filtered$number_of_crashes,pred = df_test_filtered$preds, 
                      ymax = df_test_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_test_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])
```

the model is better but still no great

we probably need more data.

using all the available data:

```{r}
full_collisioin_data <- readr::read_csv(file="dft-road-casualty-statistics-collision-1979-latest-published-year.csv") 

spec(full_collisioin_data)
glimpse(full_collisioin_data)

full_collisions_all <- stats19::format_collisions(full_collisioin_data)

glimpse(full_collisions_all)
```

```{r}
all_dates <- unique(full_collisions_all$date)
all_police_forces <- unique(full_collisions_all$police_force)

full_combinations <- expand.grid(date = all_dates, police_force = all_police_forces)

crashes_per_day_police <- full_combinations %>%
  left_join(full_collisions_all %>% group_by(date, police_force) %>% 
              summarize(number_of_crashes = n()),
            by = c("date", "police_force")) %>%
  replace_na(list(number_of_crashes = 0))

glimpse(crashes_per_day_police)
```

```{r}
crashes_per_day_police <- crashes_per_day_police %>% mutate(day_of_year = yday(date),
                                                            day = factor(wday(date)),year = year(date))
```

removing the 29 of february for dimensioality problems in the vectors below

```{r}
crashes_per_day_police <- crashes_per_day_police %>% filter(!(month(date)==2 & day(date)==29))
```

```{r}
train_years <- ceiling(seq(1979,2021,by = 1.5))
cal_years <- seq(1979,2021,3)

df_train <- crashes_per_day_police %>% filter(year %in% train_years) 
df_cal <- crashes_per_day_police %>% filter(year %in% cal_years) 
df_test <- crashes_per_day_police %>% filter(year == 2022)
```

starting with a gam and mixed effects

```{r}
gam_train <- bam(number_of_crashes ~ day + year +
                                  s(day_of_year,k = 53, bs = "cr") + 
                       s(police_force, bs = "re"), 
                data = df_train, family=quasipoisson(), method='REML')
```

```{r}
summary(gam_train)
```

```{r}
b <- getViz(gam_train)
```

```{r}
print(plot(b, allTerms = T), pages = 1)
```

1 is sunday here

```{r}
newd <- with(df_train, data.frame(day_of_year = 120,
                                  day = 1,
                                  year = 2022,
                                  police_force = levels(police_force)))

p <- predict(gam_train, newd, type = "terms", se.fit = TRUE)
re <- p[["fit"]][ , "s(police_force)"]
se <- p[["se.fit"]][ , "s(police_force)"]

data <- data.frame(police_force = levels(df_train$police_force), effect = re,std.err = se)
```

```{r}
data %>% 
ggplot(aes(effect, fct_reorder(police_force, effect))) +
  geom_vline(xintercept = 0, color = "gray50", lty = 2, linewidth = 1.2) +
  geom_errorbar(aes(
    xmin = effect - 1.96*std.err,
    xmax = effect + 1.96*std.err),width = 0.5, alpha = 0.7) +
  geom_point(size = 1,colour = "blue") + theme_minimal() + 
  labs(y = "District", x = "Random effect") + ggtitle("Random effects for each police force") + 
  theme(legend.position = "none",plot.title = element_text(size = 16,hjust = 0.5)) 
```

```{r}
check(b)
```

we now need to compute the bands in a functional case

```{r}
res_train <- cbind(df_train,preds = predict(gam_train, df_train, type = "response")) %>%
  arrange(police_force,date) %>% mutate(absdiff = abs(number_of_crashes-preds))

res_cal <- cbind(df_cal,preds = predict(gam_train, df_cal, type = "response")) %>%
  arrange(police_force,date) %>% mutate(absdiff = abs(number_of_crashes-preds))

res_test <- cbind(df_test,preds = predict(gam_train, df_test, type = "response")) %>%
  arrange(police_force,date) %>% mutate(absdiff = abs(number_of_crashes-preds))
```

```{r}
n_pol <- length(levels(crashes_per_day_police$police_force))

m = length(train_years)
l = length(cal_years)
alpha = 0.1
S = matrix(nrow = n_pol,ncol = 365)
k = rep(0,n_pol)
x = 1:365
line_integral = function(x, y) {
  dx = diff(x)
  end = length(y)
  my = (y[1:(end - 1)] + y[2:end]) / 2
  sum(dx *my)
} 

ceiling((m+1)*(1-alpha))
```

H1 != I1 now

computing the modulation functions:

```{r}
for(i in 1:n_pol){
  ad_train <- res_train %>% filter(police_force == levels(police_force)[i]) %>%
  dplyr::select(absdiff)
  a_mat <- matrix(ad_train$absdiff,byrow = F,nrow = 365)
  setcfr <- apply(a_mat,2,max)
  gamma <- sort(setcfr)[ceiling((m+1)*(1-alpha))]
  a_mat_filt <- a_mat[,setcfr<=gamma]
  ncm = apply(a_mat_filt,1,max)
  ncm_sort <- sort(ncm)
  den = line_integral(x, ncm)
  s = ncm/den
  S[i,] = s
}
```

computing the k:

```{r}
ceiling((l + 1)*(1-alpha))
```

we use the biggest value for the choice of k

```{r}
for(i in 1:n_pol){
  ad_cal <- res_cal %>% filter(police_force == levels(police_force)[i]) %>%
  dplyr::select(absdiff)
  a_mat <- matrix(ad_cal$absdiff,byrow = F,nrow = 365)
  ncs = apply(a_mat/S[i,],1,max)
  k[i] = max(ncs)
}
```

```{r}
i <- 50

df_filtered <- res_test %>% filter(police_force == levels(police_force)[i]) %>%
  arrange(day_of_year)

df_plot <- data.frame(n = df_filtered$number_of_crashes,pred = df_filtered$preds, 
                      ymax = df_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])

```

this is not great:


```{r}
i <- 6

df_filtered <- res_test %>% filter(police_force == levels(police_force)[i]) %>%
  arrange(day_of_year)

df_plot <- data.frame(n = df_filtered$number_of_crashes,pred = df_filtered$preds, 
                      ymax = df_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])

```


```{r}
i <- 30

df_filtered <- res_test %>% filter(police_force == levels(police_force)[i]) %>%
  arrange(day_of_year)

df_plot <- data.frame(n = df_filtered$number_of_crashes,pred = df_filtered$preds, 
                      ymax = df_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])

```

this is eve worse tha before:

doing it for a single police force at a time:

```{r}
gams <- list()

S = matrix(nrow = n_pol,ncol = 365)
k = rep(0,n_pol)

for(i in 1:n_pol){
  
  df_train_mid <- df_train %>% filter(police_force == levels(police_force)[i])
  df_cal_mid <- df_cal %>% filter(police_force == levels(police_force)[i])
  
  gam <- bam(number_of_crashes ~ day + year +
                                  s(day_of_year,k = 53, bs = "cr"), 
                data = df_train_mid, family=quasipoisson(), method='REML')
  
  res_train_mid <- cbind(df_train_mid,preds = predict(gam, df_train_mid, type = "response")) %>%
  arrange(date) %>% mutate(absdiff = abs(number_of_crashes-preds))

  res_cal_mid <- cbind(df_cal_mid,preds = predict(gam, df_cal_mid, type = "response")) %>%
  arrange(date) %>% mutate(absdiff = abs(number_of_crashes-preds))
  
  ad_train <- res_train_mid %>% filter(police_force == levels(police_force)[i]) %>%
  dplyr::select(absdiff)
  a_mat <- matrix(ad_train$absdiff,byrow = F,nrow = 365)
  setcfr <- apply(a_mat,2,max)
  gamma <- sort(setcfr)[ceiling((m+1)*(1-alpha))]
  a_mat_filt <- a_mat[,setcfr<=gamma]
  ncm = apply(a_mat_filt,1,max)
  ncm_sort <- sort(ncm)
  den = line_integral(x, ncm)
  s = ncm/den
  S[i,] = s
  
  ad_cal <- res_cal_mid %>% filter(police_force == levels(police_force)[i]) %>%
  dplyr::select(absdiff)
  a_mat <- matrix(ad_cal$absdiff,byrow = F,nrow = 365)
  ncs = apply(a_mat/S[i,],1,max)
  k[i] = max(ncs)
  
  gams[[i]] <- gam
}
```

midlands

```{r}
i <- 50
df_test_filtered <- df_test %>% filter(police_force == levels(police_force)[i])
preds <- predict(gams[[i]], df_test_filtered, type = "response")
df_test_filtered  <- cbind(df_test_filtered,preds) %>% arrange(day_of_year)
```

```{r}
df_plot <- data.frame(n = df_test_filtered$number_of_crashes,pred = df_test_filtered$preds, 
                      ymax = df_test_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_test_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])
```

London:

```{r}
i <- 30
df_test_filtered <- df_test %>% filter(police_force == levels(police_force)[i])
preds <- predict(gams[[i]], df_test_filtered, type = "response")
df_test_filtered  <- cbind(df_test_filtered,preds) %>% arrange(day_of_year)
```

```{r}
df_plot <- data.frame(n = df_test_filtered$number_of_crashes,pred = df_test_filtered$preds, 
                      ymax = df_test_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_test_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])
```

the model is still ot graet

