---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "Expolratory data analysis collisions data"
author:
    - "Valeria Iapaolo"
    - "Oswald Jesus Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: true
    pdf_document:
        toc: true
        toc_depth: 3
        number_section: true
date: "2023-11-10"
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center',
    #fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r, include=FALSE}
library(tidyverse)
library(lubridate)
library(patchwork)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/full_vehicles.RData")
```

```{r}
glimpse(full_vehicles)
```

## vehicle type:

```{r}
full_vehicles %>% ggplot(aes(vehicle_type)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

there is a large number of accidents with bikes and motorcyle, which could be interesting to highlight

## towing_and_articulation:

```{r}
full_vehicles %>% ggplot(aes(towing_and_articulation)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

probably not very usefull

## vehicle_manoeuvre

```{r}
full_vehicles %>% ggplot(aes(vehicle_manoeuvre)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

a lot more turning right than left, makes sense sicne they drive on the opposite side of the road

## vehicle_direction_from:

```{r}
full_vehicles %>% ggplot(aes(vehicle_direction_from)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

## vehicle_direction_to:

```{r}
full_vehicles %>% ggplot(aes(vehicle_direction_to)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

don't know why south east is dominant

## vehicle type:

```{r}
full_vehicles %>% ggplot(aes(vehicle_type)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

## vehicle_location_restricted_lane:

```{r}
full_vehicles %>% ggplot(aes(vehicle_location_restricted_lane)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 60))
```

## junction_location:

```{r}
full_vehicles %>% ggplot(aes(junction_location)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 60))
```

## skidding_and_overturning:

```{r}
full_vehicles %>% ggplot(aes(skidding_and_overturning)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

## hit_object_in_carriageway:

```{r}
full_vehicles %>% ggplot(aes(hit_object_in_carriageway)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

## hit_object_off_carriageway:

```{r}
full_vehicles %>% ggplot(aes(hit_object_off_carriageway)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

## vehicle_leaving_carriageway:

```{r}
full_vehicles %>% ggplot(aes(vehicle_leaving_carriageway)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```
## first_point_of_impact:

```{r}
full_vehicles %>% ggplot(aes(first_point_of_impact)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

i thik this could be useful, we need to match with severity by joining

## vehicle_left_hand_drive:

```{r}
full_vehicles %>% ggplot(aes(vehicle_left_hand_drive)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

probably useless

## journey_purpose_of_driver:

```{r}
full_vehicles %>% ggplot(aes(journey_purpose_of_driver)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

## sex_of_driver:

```{r}
full_vehicles %>% ggplot(aes(sex_of_driver)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

## age_band_of_driver:

```{r}
full_vehicles %>% ggplot(aes(age_band_of_driver)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

the peak is not at the young, we would need to factor in the distrivution of the population

## propulsion_code:

```{r}
full_vehicles %>% ggplot(aes(propulsion_code)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

## age_of_vehicle:

```{r}
full_vehicles %>% ggplot(aes(age_of_vehicle)) + geom_histogram(bins = 20) + xlim(c(1,20))
```

