---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "Expolratory data analysis collisions data"
author:
    - "Valeria Iapaolo"
    - "Oswald Jesus Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: true
    pdf_document:
        toc: true
        toc_depth: 3
        number_section: true
date: "2023-11-10"
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center',
    #fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r, include=FALSE}
library(tidyverse)
library(lubridate)
library(patchwork)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/full_collisions.RData")
```

```{r}
glimpse(full_collisions)
```

doing some EDA with a focus on severity:

## severity

```{r}
full_collisions %>% ggplot(aes(accident_severity)) + geom_bar()
```
## year

```{r}
full_collisions %>% ggplot(aes(accident_year)) + geom_bar(aes(fill = accident_severity))
```

fixing the scale to see the proportions:

```{r}
full_collisions %>% ggplot(aes(y = accident_year, fill = accident_severity)) + 
  geom_bar(position = "fill")
```

the composition seems to change over the years, more serious than slight

## time

```{r}
full_collisions %>% ggplot(aes(time,color = accident_severity)) + geom_freqpoly(bins = 48)
```

```{r}
full_collisions %>%  filter(accident_severity == "Fatal") %>% ggplot(aes(time,color = accident_severity)) + geom_freqpoly(bins = 48)
```

same pattern for this.

```{r}
full_collisions %>% ggplot(aes(x = time, fill = accident_severity)) + 
  geom_histogram(bins = 48,position = "fill")
```

this could be interesting as well, in the night the severity increases.

## day of the week

```{r}
full_collisions$day_of_week <- factor(full_collisions$day_of_week,levels = c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"),ordered = T)

full_collisions %>% ggplot(aes(day_of_week)) + geom_bar(aes(fill = accident_severity))
```

```{r}
full_collisions %>% ggplot(aes(x = day_of_week, fill = accident_severity)) + 
  geom_bar(position = "fill")
```

the severity changes

## date

```{r}
full_collisions %>% mutate(n_accidents = n(),day = yday(date)) %>%  ggplot(aes(day,color = as.factor(accident_year))) + geom_freqpoly(bins = 366) + theme(legend.position = "none")
```

we can see the covid years:

clearly:

```{r}
full_collisions %>% filter(accident_year > 2018) %>% mutate(n_accidents = n(),day = yday(date)) %>%  ggplot(aes(day,color = as.factor(accident_year))) + geom_freqpoly(bins = 366) + theme(legend.position = "top")
```

we can try to look for a trend in the day of the month:

```{r}
full_collisions %>% mutate(n_accidents = n(),day = mday(date)) %>%  ggplot(aes(day,color = as.factor(accident_year))) + geom_freqpoly(bins = 31) + theme(legend.position = "none")
```
nothing

## number of vehicles:

```{r}
full_collisions %>% ggplot(aes(number_of_vehicles)) + geom_bar(aes(fill = accident_severity)) + xlim(c(1,10))
```


```{r}
full_collisions %>% ggplot(aes(x = number_of_vehicles, fill = accident_severity)) + 
  geom_bar(position = "fill") + xlim(c(1,10))
```


## number of casuallties:

```{r}
full_collisions %>% ggplot(aes(number_of_casualties)) + geom_bar(aes(fill = accident_severity)) + xlim(c(1,10))
```


```{r}
full_collisions %>% ggplot(aes(x = number_of_casualties, fill = accident_severity)) + 
  geom_bar(position = "fill") + xlim(c(1,10))
```

## first road class:

```{r}
full_collisions %>% ggplot(aes(first_road_class)) + geom_bar(aes(fill = accident_severity)) 
```

```{r}
full_collisions %>% ggplot(aes(x = first_road_class, fill = accident_severity)) + 
  geom_bar(position = "fill")
```

## second roadd class

```{r}
full_collisions %>% ggplot(aes(second_road_class)) + geom_bar(aes(fill = accident_severity)) + scale_x_discrete(guide = guide_axis(angle = 90))
```

```{r}
full_collisions %>% ggplot(aes(x = second_road_class, fill = accident_severity)) + 
  geom_bar(position = "fill") + scale_x_discrete(guide = guide_axis(angle = 90))
```

## road type

```{r}
full_collisions %>% ggplot(aes(road_type)) + geom_bar(aes(fill = accident_severity)) + scale_x_discrete(guide = guide_axis(angle = 90))
```

```{r}
full_collisions %>% ggplot(aes(x = road_type, fill = accident_severity)) + 
  geom_bar(position = "fill") + scale_x_discrete(guide = guide_axis(angle = 90))
```

## speed limit

```{r}
full_collisions %>% ggplot(aes(as.factor(speed_limit))) + geom_bar(aes(fill = accident_severity)) 
```

```{r}
full_collisions %>% ggplot(aes(x = as.factor(speed_limit), fill = accident_severity)) + 
  geom_bar(position = "fill") 
```

## junction detail

```{r}
full_collisions %>% ggplot(aes(junction_detail)) + geom_bar(aes(fill = accident_severity)) + scale_x_discrete(guide = guide_axis(angle = 90))
```

```{r}
full_collisions %>% ggplot(aes(x = junction_detail, fill = accident_severity)) + 
  geom_bar(position = "fill") + scale_x_discrete(guide = guide_axis(angle = 90))
```

## junction control

```{r}
full_collisions %>% ggplot(aes(junction_control)) + geom_bar(aes(fill = accident_severity)) + scale_x_discrete(guide = guide_axis(angle = 90))
```

```{r}
full_collisions %>% ggplot(aes(x = junction_control, fill = accident_severity)) + 
  geom_bar(position = "fill") + scale_x_discrete(guide = guide_axis(angle = 90))
```

## pedestrian_crossing_human_control

```{r}
p1 <- full_collisions %>% ggplot(aes(pedestrian_crossing_human_control)) + 
  geom_bar(aes(fill = accident_severity)) + scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme(legend.position = "none")

p2 <- full_collisions %>% ggplot(aes(x = pedestrian_crossing_human_control, fill = accident_severity)) + 
  geom_bar(position = "fill") + scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme(legend.position = "none")
p1 + p2
```

## pedestrian_crossing_physical_facilities

```{r}
p1 <- full_collisions %>% ggplot(aes(pedestrian_crossing_human_control)) + 
  geom_bar(aes(fill = accident_severity)) + scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme(legend.position = "none")

p2 <- full_collisions %>% ggplot(aes(x = pedestrian_crossing_human_control, fill = accident_severity)) + 
  geom_bar(position = "fill") + scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme(legend.position = "none")
p1 + p2
```

## light_conditions

```{r}
p1 <- full_collisions %>% ggplot(aes(light_conditions)) + 
  geom_bar(aes(fill = accident_severity)) + scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme(legend.position = "none")

p2 <- full_collisions %>% ggplot(aes(x = light_conditions, fill = accident_severity)) + 
  geom_bar(position = "fill") + scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme(legend.position = "none")
p1 + p2
```

## weather_conditions

```{r}
p1 <- full_collisions %>% ggplot(aes(weather_conditions)) + 
  geom_bar(aes(fill = accident_severity)) + scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme(legend.position = "none")

p2 <- full_collisions %>% ggplot(aes(x = weather_conditions, fill = accident_severity)) + 
  geom_bar(position = "fill") + scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme(legend.position = "none")
p1 + p2
```

## special_conditions_at_site

```{r}
p1 <- full_collisions %>% ggplot(aes(special_conditions_at_site)) + 
  geom_bar(aes(fill = accident_severity)) + scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme(legend.position = "none")

p2 <- full_collisions %>% ggplot(aes(x = special_conditions_at_site, fill = accident_severity)) + 
  geom_bar(position = "fill") + scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme(legend.position = "none")
p1 + p2
```

## carriageway_hazards

```{r}
p1 <- full_collisions %>% ggplot(aes(carriageway_hazards)) + 
  geom_bar(aes(fill = accident_severity)) + scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme(legend.position = "none")

p2 <- full_collisions %>% ggplot(aes(x = carriageway_hazards, fill = accident_severity)) + 
  geom_bar(position = "fill") + scale_x_discrete(guide = guide_axis(angle = 60)) +
  theme(legend.position = "none")
p1 + p2
```

## did_police_officer_attend_scene_of_accident

```{r}
p1 <- full_collisions %>% ggplot(aes(did_police_officer_attend_scene_of_accident)) + 
  geom_bar(aes(fill = accident_severity)) + scale_x_discrete(guide = guide_axis(angle = 45)) +
  theme(legend.position = "none")

p2 <- full_collisions %>% ggplot(aes(x = did_police_officer_attend_scene_of_accident, fill = accident_severity)) + 
  geom_bar(position = "fill") + scale_x_discrete(guide = guide_axis(angle = 45)) +
  theme(legend.position = "none")
p1 + p2
```

## trunk_road_flag

```{r}
p1 <- full_collisions %>% ggplot(aes(trunk_road_flag)) + 
  geom_bar(aes(fill = accident_severity)) + scale_x_discrete(guide = guide_axis(angle = 45)) +
  theme(legend.position = "none")

p2 <- full_collisions %>% ggplot(aes(x = trunk_road_flag, fill = accident_severity)) + 
  geom_bar(position = "fill") + scale_x_discrete(guide = guide_axis(angle = 45)) +
  theme(legend.position = "none")
p1 + p2
```



