---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "Expolratory data analysis collisions data"
author:
    - "Valeria Iapaolo"
    - "Oswald Jesus Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
output:
    html_document:
        toc: true
        toc_float: true
        number_sections: true
    pdf_document:
        toc: true
        toc_depth: 3
        number_section: true
date: "2023-11-10"
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center',
    #fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r, include=FALSE}
library(tidyverse)
library(lubridate)
library(patchwork)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/full_casualties.RData")
```

```{r}
glimpse(full_casualties)
```

eda with a focus on severity:

## casualty_severity:

```{r}
full_casualties %>% ggplot(aes(casualty_severity)) + geom_bar() + 
  scale_x_discrete(guide = guide_axis(angle = 90))
```

## accident_year:

```{r}
full_casualties %>% ggplot(aes(as.factor(accident_year))) + geom_bar(aes(fill = casualty_severity)) + 
  scale_x_discrete(guide = guide_axis(angle = 15)) 
```

fixing the scale to see the proportions:

```{r}
full_casualties %>% ggplot(aes(y = accident_year, fill = casualty_severity)) + 
  geom_bar(position = "fill")
```

this is a downward trend

## casualty_class:

```{r}
p1 <- full_casualties %>% ggplot(aes(casualty_class)) + geom_bar(aes(fill = casualty_severity)) + theme(legend.position = "none")
p2 <- full_casualties %>% ggplot(aes(x = casualty_class, fill = casualty_severity)) + 
  geom_bar(position = "fill") + theme(legend.position = "none")
p1 + p2
```


## sex_of_casualty:

```{r}
p1 <- full_casualties %>% ggplot(aes(sex_of_casualty)) + geom_bar(aes(fill = casualty_severity)) + theme(legend.position = "none") + scale_x_discrete(guide = guide_axis(angle = 15))
p2 <- full_casualties %>% ggplot(aes(x = sex_of_casualty, fill = casualty_severity)) + 
  geom_bar(position = "fill") + theme(legend.position = "none") + 
  scale_x_discrete(guide = guide_axis(angle = 15))
p1 + p2
```

## age_band_of_casualty:

```{r}
full_casualties %>% ggplot(aes(age_band_of_casualty)) + geom_bar(aes(fill = casualty_severity)) + 
  scale_x_discrete(guide = guide_axis(angle = 15))
```

```{r}
full_casualties %>% ggplot(aes(age_band_of_casualty)) + 
  geom_bar(aes(fill = casualty_severity),position = "fill") + 
  scale_x_discrete(guide = guide_axis(angle = 15))
```


## casualty_class:

```{r}
p1 <- full_casualties %>% ggplot(aes(casualty_class)) + geom_bar(aes(fill = casualty_severity)) + theme(legend.position = "none") + scale_x_discrete(guide = guide_axis(angle = 15))
p2 <- full_casualties %>% ggplot(aes(x = casualty_class, fill = casualty_severity)) + 
  geom_bar(position = "fill") + theme(legend.position = "none") + 
  scale_x_discrete(guide = guide_axis(angle = 15))
p1 + p2
```

## pedestrian_location:

```{r}
full_casualties %>%  filter(pedestrian_location != "Not a Pedestrian") %>% ggplot(aes(pedestrian_location)) + geom_bar(aes(fill = casualty_severity)) + 
  scale_x_discrete(guide = guide_axis(angle = 70)) + theme(legend.position = "none")
```

```{r}
full_casualties %>%  filter(pedestrian_location != "Not a Pedestrian") %>% ggplot(aes(pedestrian_location)) + geom_bar(aes(fill = casualty_severity),position = "fill") + 
  scale_x_discrete(guide = guide_axis(angle = 70)) + theme(legend.position = "none",axis.text.x=element_blank())
```

## pedestrian_movement:

```{r}
levels(full_casualties$pedestrian_movement)
```


```{r}
full_casualties %>%  filter(pedestrian_movement != "Not a Pedestrian") %>% ggplot(aes(pedestrian_movement)) + geom_bar(aes(fill = casualty_severity)) + 
  scale_x_discrete(guide = guide_axis(angle = 70)) + theme(legend.position = "none",axis.text.x=element_blank())
```

```{r}
full_casualties %>%  filter(pedestrian_movement != "Not a Pedestrian") %>% ggplot(aes(pedestrian_movement)) + geom_bar(aes(fill = casualty_severity),position = "fill") + theme(legend.position = "none",axis.text.x=element_blank())
```

## car_passenger:

```{r}
full_casualties %>% ggplot(aes(car_passenger)) + geom_bar(aes(fill = casualty_severity)) + 
  scale_x_discrete(guide = guide_axis(angle =45))
```

this could be interesting for the car manufacturer

## bus_or_coach_passenger:

```{r}
full_casualties %>% filter(bus_or_coach_passenger != "Not a bus or coach passenger") %>% ggplot(aes(bus_or_coach_passenger)) + geom_bar(aes(fill = casualty_severity)) + 
  scale_x_discrete(guide = guide_axis(angle = 15))
```

```{r}
full_casualties %>% ggplot(aes(bus_or_coach_passenger)) + geom_bar(aes(fill = casualty_severity),position = "fill") + 
  scale_x_discrete(guide = guide_axis(angle = 15)) + theme(legend.position = "none")
```

## casualty_type:

```{r}
full_casualties %>% ggplot(aes(casualty_type)) + geom_bar(aes(fill = casualty_severity)) + 
  scale_x_discrete(guide = guide_axis(angle = 60)) + theme(legend.position = "none")
```

```{r}
full_casualties %>% ggplot(aes(casualty_type)) + geom_bar(aes(fill = casualty_severity),position = "fill")  + theme(legend.position = "none",axis.text.x=element_blank())
```

this should be investigated, we need to make more informad categories

## casualty_home_area_type

```{r}
full_casualties %>% ggplot(aes(casualty_home_area_type)) + geom_bar(aes(fill = casualty_severity)) + 
  scale_x_discrete(guide = guide_axis(angle = 15)) + theme(legend.position = "none")
```
