---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "Functional permutation testing weekly data"
author:
    - "Valeria Iapaolo"
    - "Oswaldo Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
    
output: html_notebook
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center',
    #fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r libraries inclusions, include=FALSE}
library(tidyverse)
library(roahd)
library(fdacluster)
library(fdatest)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/functional/df_week.RData")
```

```{r}
fd_week <- fData(1:7,as_tibble(df_week[,3:9]))
```

```{r}
plot(fd_week,lwd = 1,xlab = "day",ylab = "dayly number of crashes",main = "dayly crashes in each week")
```

we recover the 2 clusters:

```{r}
days <- 1:7
n <- fd_week$N
x <- t(matrix(rep(days,n),7,n))
y <- as.matrix(df_week[,3:9])
```

```{r}
k <- 2
system.time(
fdakma0der <- fdakmeans(x = x,y = y, n_clusters = k,
  seeds = c(832,197),
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 12
)
)
```

```{r}
autoplot(fdakma0der,type = "amplitude")
```

```{r}
matplot(t(fdakma0der$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:5,lwd = 3)
```

```{r}
table(df_week[fdakma0der$memberships==1,2])
```

```{r}
table(df_week[fdakma0der$memberships==2,2])
```

I start with a permutation test on the global curves:

```{r}
i1 <- which(fdakma0der$memberships==1)
n1 <- length(i1)

i2 <- which(fdakma0der$memberships==2)
n2 <- length(i2)
```

```{r}
seed=7122023
B=10000
```

start by comparing group 1 and 2:

```{r}
aug_df=y
n= n1 + n2 

meandiff=(colMeans(y[i1,])-colMeans(y[i2,]))
plot(meandiff,type = 'l')
T0=sum(meandiff^2)
T0
```

And, Knowing that under $H_0$ the two groups of curves are IID, my likelihood-invariant permutation scheme is of course label permutation, so...

```{r}
T0_perm=numeric(B)

for(perm in 1:B){
  permutation <- sample(n)
  df_perm=aug_df[permutation,]
  perm_1 = df_perm[1:n1,] 
  perm_2 = df_perm[(n1+1):n,] 
  T0_perm[perm]=sum(((colMeans(perm_1)-colMeans(perm_2)))^2)
}

sum(T0_perm >= T0)/B

hist(T0_perm,xlim = c(0,2*T0))
abline(v=T0,col='green')
```

this are statistically different.

we now move to a local test.

we use the interval wise procedure, so we control the interval-wise error rate

between clusters 1 an 2

```{r}
tst=IWT2(y[i1,],y[i2,],B=10000)
plot(tst)
```

as expected the difference is in the working days.