---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "Functional permutation testing dayly data"
author:
    - "Valeria Iapaolo"
    - "Oswaldo Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
    
output: html_notebook
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center',
    #fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r libraries inclusions, include=FALSE}
library(tidyverse)
library(roahd)
library(fdacluster)
library(fdatest)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/functional/df_hour.RData")
```

```{r}
fd_day <- fData(1:24,as_tibble(df_hour[,2:25]))
```

```{r}
plot(fd_day,lwd = 1,xlab = "day",ylab = "dayly number of crashes",main = "hourly crashes in each week")
```

```{r}
hours <- 1:24
n <- fd_day$N
x <- t(matrix(rep(hours,n),24,n))
y <- as.matrix(df_hour[,2:25])
```

```{r}
k <- 2
system.time(
fdakma0der <- fdakmeans(x = x,y = y, n_clusters = k,
  seeds = c(4620,6524),
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 20L
)
)
```

```{r}
autoplot(fdakma0der)
```

```{r}
matplot(t(fdakma0der$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:k,lwd = 3)
```

```{r}
i1 <- which(fdakma0der$memberships==1)
i2 <- which(fdakma0der$memberships==2)

n <- dim(df_hour)[1]
indexes_plot <- c(i1[sample(length(i1),75)],i2[sample(length(i2),75)])
nplots <- 150
unicurves <- fdakma0der$original_curves[indexes_plot, 1, ]
grids <- matrix(rep(1:24,nplots),nrow = nplots, ncol = 24,byrow = TRUE)
memberships <- fdakma0der$memberships[indexes_plot]

wrangled_data <- tibble::tibble(
      grid = purrr::array_tree(grids, margin = 1),
      value = purrr::array_tree(unicurves, margin = 1),
      membership = memberships,
      curve_id = 1:nplots
    ) |>
      tidyr::unnest(cols = c(.data$grid, .data$value)) |> 
  dplyr::bind_rows() |>
    dplyr::mutate(
      membership = as.factor(.data$membership),
      curve_id = as.factor(.data$curve_id),
    )

```

```{r}
P1 <- wrangled_data |>
      ggplot2::ggplot(ggplot2::aes(
        x = .data$grid,
        y = .data$value,
        color = .data$membership,
        group = .data$curve_id
      )) +
      ggplot2::geom_line() +
      ggplot2::facet_wrap(ggplot2::vars(.data$membership)
      ) +
      ggplot2::labs(
        x = "hour",
        y = "number of collisions per hour"
      ) +
      ggplot2::theme(legend.position = "none") 
P1
```

```{r}
P2 <- wrangled_data |>
      ggplot2::ggplot(ggplot2::aes(
        x = .data$grid,
        y = .data$value,
        color = .data$membership,
        group = .data$curve_id
      )) +
      ggplot2::geom_line() +
      ggplot2::labs(
        x = "hour",
        y = "number of collisions per hour"
      ) +
      ggplot2::theme(legend.position = "none")

P2

```


this is a clear distinction between working days and holydays.

I start with a permutation test on the global curves:

```{r}
i1 <- which(fdakma0der$memberships==1)
n1 <- length(i1)

i2 <- which(fdakma0der$memberships==2)
n2 <- length(i2)
```

```{r}
seed=7122023
B=10000
```

start by comparing group 1 and 2:

```{r}
aug_df=y
n= n1 + n2 

meandiff=(colMeans(y[i1,])-colMeans(y[i2,]))
plot(meandiff,type = 'l')
T0=sum(meandiff^2)
T0
```

```{r}
T0_perm=numeric(B)

for(perm in 1:B){
  permutation <- sample(n)
  df_perm=aug_df[permutation,]
  perm_1 = df_perm[1:n1,] 
  perm_2 = df_perm[(n1+1):n,] 
  T0_perm[perm]=sum(((colMeans(perm_1)-colMeans(perm_2)))^2)
}

sum(T0_perm >= T0)/B

hist(T0_perm,xlim = c(0,1.2*T0))
abline(v=T0,col='green')
```

this are statistically different.

we now move to a local test.

we use the interval wise procedure, so we control the interval-wise error rate

between clusters 1 an 2

```{r}
tst=IWT2(y[i1,],y[i2,],B=1000)
plot(tst)
```

as expected the difference is both in the evening and in the morning.