---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "Functional clustering monthly data"
author:
    - "Valeria Iapaolo"
    - "Oswaldo Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
    
output: html_notebook
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center',
    #fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r libraries inclusions, include=FALSE}
library(tidyverse)
library(roahd)
#library(fda)
library(fdacluster)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/functional/df_month.RData")
```

first we start with outlier detection in the functional case

```{r}
fd_month <- fData(1:31,as_tibble(df_month[,3:33]))
```

```{r}
plot(fd_month,lwd = 3,xlab = "day",ylab = "dayly number of crashes",main = "dayly crashes in each month")
```
functional bagplot:

```{r}
year_fbagplot <- fbplot(fd_month, main="Magnitude outliers monthly data")
```

the default F is:

```{r}
year_fbagplot$Fvalue
```

no outliers found

```{r}
df_month[year_fbagplot$ID_outliers,1:2]
```

we need to remove the day 366 since it causes problems in the iterations to find the F:

```{r}
fd_month2 <- fData(1:28,df_month[,3:30])
```

```{r}
year_fbagplot2 <- fbplot(fd_month, main="Magnitude outliers yearly data",
                                  adjust = list( N_trials = 20,trial_size = fd_month2$N,
                                                 VERBOSE = TRUE ))
```

the chosen F value is:

```{r}
year_fbagplot2$Fvalue
```

the outlying years are:

```{r}
df_month[year_fbagplot2$ID_outliers,1:2]
```

nothing changedjust the first year are seen as outliers in this case

outiliergram:

```{r}
invisible(out_month <- outliergram(fd_month,adjust = F,lwd = 3,display = F))
```

the found outliers are:

```{r}
df_month[out_month$ID_outliers,1:2]
```

this are all in december.

the plot of the original function is not working.

adjusting the F:

```{r}
out_month <- outliergram(fd_month,lwd = 5,adjust = list( N_trials = 20,trial_size = 8*fd_month$N,
                                                 VERBOSE = TRUE ),display = FALSE)
```

```{r}
out_month$Fvalue
```

nothing changed, same outliers detected.

```{r}
df_month[out_month$ID_outliers,1:2]
```

plotting in the old way.

```{r}
par(mfrow=c(1,2))
plot(fd_month[out_month$ID_outliers,],lwd = 3,main = "outliers",col = 2)
plot(fd_month[-out_month$ID_outliers,],lwd = 3,main = "non outliers",col = 3)
```

there is a clear pattern in the data due to the christmas period.

doing some clustering:

let's run a comparison of the methods:

```{r}
days <- 1:31
n <- fd_month$N
x <- t(matrix(rep(days,n),31,n))
y <- as.matrix(df_month[,3:33])
```

```{r}
k <- 3
system.time(
fdakma0der_shift <- fdakmeans(x = x,y = y, n_clusters = k,
  seeding_strategy = "kmeans++",
  warping_class = "shift",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = TRUE
)
)
```

```{r}
autoplot(fdakma0der_shift,type = "amplitude")
```

```{r}
autoplot(fdakma0der_shift,type = "phase")
```

```{r}
diagnostic_plot(fdakma0der_shift)
```

the negative silouettes are not good

```{r}
n_sub <- 50
sub_id <- sample(1:n,n_sub,replace = FALSE)
x_sub <- t(matrix(rep(days,n_sub),31,n_sub))
y_sub <- as.matrix(df_month[sub_id,3:33])

system.time(invisible(comparison_kmeans <- compare_caps(
  x_sub,
  y_sub,
  n_clusters = 2:5,
  metric = "pearson",
  clustering_method = "kmeans",
  warping_class = "shift",
  centroid_type = "mean",
  cluster_on_phase = FALSE
    )
  )
)
```

```{r}
plot(comparison_kmeans, validation_criterion = "wss", what = "mean",lwd = 3)
```

```{r}
plot(comparison_kmeans, validation_criterion = "wss", what = "distribution")
```

```{r}
plot(comparison_kmeans, validation_criterion = "silhouette", what = "mean")
```

```{r}
plot(comparison_kmeans, validation_criterion = "silhouette", what = "distribution")
```


```{r}
n_sub <- 50
sub_id <- sample(1:n,n_sub,replace = FALSE)
x_sub <- t(matrix(rep(days,n_sub),31,n_sub))
y_sub <- as.matrix(df_month[sub_id,3:33])

system.time(invisible(comparison_kmeans <- compare_caps(
  x_sub,
  y_sub,
  n_clusters = 2:5,
  metric = "pearson",
  clustering_method = "kmeans",
  warping_class = "none",
  centroid_type = "mean",
  cluster_on_phase = FALSE
    )
  )
)
```

```{r}
plot(comparison_kmeans, validation_criterion = "wss", what = "mean",lwd = 3)
```

```{r}
plot(comparison_kmeans, validation_criterion = "wss", what = "distribution")
```

```{r}
plot(comparison_kmeans, validation_criterion = "silhouette", what = "mean")
```

```{r}
plot(comparison_kmeans, validation_criterion = "silhouette", what = "distribution")
```

this does not seem conclusive

let's do an experiment on the whole dataset for k = 4:

```{r}
k <- 4
seeds <- c(184,193,113,50) # from a previous experiment using kmeans++
system.time(
fdakma0der_shift <- fdakmeans(x = x,y = y, n_clusters = k,
  seeds = seeds,
  warping_class = "shift",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F
)
)
```

```{r}
autoplot(fdakma0der_shift,type = "amplitude")
```

```{r}
autoplot(fdakma0der_shift,type = "phase")
```


```{r}
matplot(t(fdakma0der_shift$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:4)
```

we have 2 curves that are similar (blue and green)

we can look for each of the clusters the month:

cluster 1:

this has more of a baseline distrtibution:
```{r}
table(df_month[fdakma0der_shift$memberships==1,2])
```

cluster 2:

this has most of the december months, since low number at the end

```{r}
table(df_month[fdakma0der_shift$memberships==2,2])
```

cluster 3:

this has most of the summer months

```{r}
table(df_month[fdakma0der_shift$memberships==3,2])
```

cluster 4:

this has most of the months of january, since low at the beginning of the month 

```{r}
table(df_month[fdakma0der_shift$memberships==4,2])
```

another possibility is to allign all of the curves befor the clustering and then do the clustering:

```{r}

fdakma0der_shift <- fdakmeans(x = x,y = y, n_clusters = 1,
  seeds = 1,
  warping_class = "shift",
  metric = "pearson",
  centroid_type = "medoid",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F
)

```

```{r}
autoplot(fdakma0der_shift,type = "amplitude")
```

```{r}
autoplot(fdakma0der_shift,type = "phase")
```

we recover the curves in the following way:

```{r}
alligned_x <- fdakma0der_shift$grids[1,]
alligned_y <- fdakma0der_shift$aligned_curves[,1,]
```

we can cluster this with no allignment:

```{r}
k <- 4
seeds <- c(184,193,113,50) # from a previous experiment using kmeans++
system.time(
fdakma0der_shift <- fdakmeans(x = alligned_x,y = alligned_y, n_clusters = k,
  seeds = seeds,
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = T
)
)
```

```{r}
autoplot(fdakma0der_shift,type = "amplitude")
```

```{r}
diagnostic_plot(fdakma0der_shift)
```

this is still not great:

the dominating behaviour is at the end of the month:

```{r}
matplot(t(fdakma0der_shift$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:4)
```

we have 2 curves that are similar (blue and green)

we can look for each of the clusters the month:

cluster 1:

this has more of a baseline distrtibution of the work months:
```{r}
table(df_month[fdakma0der_shift$memberships==1,2])
```

cluster 2:

this has most of the december months, since low number at the end

```{r}
table(df_month[fdakma0der_shift$memberships==2,2])
```

cluster 3:

this has most of the summer months

```{r}
table(df_month[fdakma0der_shift$memberships==3,2])
```

cluster 4:

this has most of the months of february, since low at the en of the month

```{r}
table(df_month[fdakma0der_shift$memberships==4,2])
```

let's try to increase the number of clusters and then possibly joining them later:

```{r}
k <- 10
system.time(
fdakma0der_shift <- fdakmeans(x = alligned_x,y = alligned_y, n_clusters = k,
  seeding_strategy = "kmeans++",
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F
)
)
```

```{r}
autoplot(fdakma0der_shift,type = "amplitude")
```

```{r}
matplot(t(fdakma0der_shift$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:k,lwd = 3)
```

we can see that we have some outliers:

cluster 6 an 9 are outliers

```{r}
labels <- fdakma0der_shift$memberships
```

```{r}
df_month[fdakma0der_shift$memberships==6,1:2]
```
this is the covid year

```{r}
df_month[fdakma0der_shift$memberships==9,1:2]
```
this has a low spike arroun day 10

the other 2 small clusters are 3 and 8

```{r}
df_month[fdakma0der_shift$memberships %in% c(3,8),1:2]
```


this are just february.

we can take a look at the others:

```{r}
matplot(t(fdakma0der_shift$center_curves[,1,])[,-c(3,6,8,9)],type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:6,lwd = 3)
```
the difference is just in the beginnig and at the end:

cluster 1:
```{r}
table(df_month[fdakma0der_shift$memberships==1,2])
```

decembers

cluster 2:

```{r}
table(df_month[fdakma0der_shift$memberships==2,2])
```

cluster 4:

this has most of the summer months

```{r}
table(df_month[fdakma0der_shift$memberships==4,2])
```

cluster 5:

this has most of the months of febrayry, since low at the beginning of the month 

```{r}
table(df_month[fdakma0der_shift$memberships==5,2])
```

custer 7:

```{r}
table(df_month[fdakma0der_shift$memberships==7,2])
```


cluster 10:
```{r}
table(df_month[fdakma0der_shift$memberships==10,2])
```

this has january, low at the beginning

this can be useful to reduce the clusters to:

2 outliers

3, 5 and 8 joined in febrayry

10 for january

4 for summer 

the rest together.

we could try using hclust:

```{r}
system.time(
  hclustres <- fdahclust(
  x = alligned_x,
  y = alligned_y,
  n_clusters = 4L,
  warping_class = "none",
  centroid_type = "mean",
  metric = "pearson",
  linkage_criterion = "complete",
  cluster_on_phase = FALSE,
  use_verbose = TRUE,
  warping_options = c(0.15, 0.15),
  maximum_number_of_iterations = 100L,
  number_of_threads = 12L,
  parallel_method = 0L,
  distance_relative_tolerance = 0.001,
  use_fence = FALSE,
  check_total_dissimilarity = TRUE,
  compute_overall_center = FALSE
)
)
```

```{r}
autoplot(hclustres)
```


```{r}
diagnostic_plot(hclustres)
```

this is still not great:

the dominating behaviour is at the end of the month:

```{r}
table(hclustres$memberships)
```


```{r}
matplot(t(hclustres$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:4,lwd = 3)
```

clsuter 1 winter an may

```{r}
table(df_month[hclustres$memberships==1,2])
```

cluster 2 ferb

```{r}
table(df_month[hclustres$memberships==2,2])
```

cluster 3 summer and spring :
```{r}
table(df_month[hclustres$memberships==3,2])
```

cluster 4  feb

```{r}
table(df_month[hclustres$memberships==4,2])
```

we need to select k, we can do a comparison with no allignlemt:

```{r}
n_sub <- 50
sub_id <- sample(1:n,n_sub,replace = FALSE)
x_sub <- t(matrix(rep(alligned_x,n_sub),31,n_sub))
y_sub <- as.matrix(alligned_y[sub_id,])

system.time(invisible(comparison_kmeans <- compare_caps(
  x_sub,
  y_sub,
  n_clusters = 3:6,
  metric = "pearson",
  clustering_method = "kmeans",
  warping_class = "none",
  centroid_type = "mean",
  cluster_on_phase = FALSE
    )
  )
)
```

```{r}
plot(comparison_kmeans, validation_criterion = "wss", what = "mean",lwd = 3)
```

```{r}
plot(comparison_kmeans, validation_criterion = "wss", what = "distribution")
```

```{r}
plot(comparison_kmeans, validation_criterion = "silhouette", what = "mean")
```

```{r}
plot(comparison_kmeans, validation_criterion = "silhouette", what = "distribution")
```

this is suggesting 3 clusters:

```{r}
k <- 3
system.time(
fdakma0der_shift <- fdakmeans(x = alligned_x,y = alligned_y, n_clusters = k,
  seeding_strategy = "kmeans++",
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 10L
)
)
```

```{r}
autoplot(fdakma0der_shift,type = "amplitude")
```

```{r}
matplot(t(fdakma0der_shift$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:3,lwd = 3)
```


cluster 1:
```{r}
table(df_month[fdakma0der_shift$memberships==1,2])
```


cluster 2:

```{r}
table(df_month[fdakma0der_shift$memberships==2,2])
```

cluster 3:

february
```{r}
table(df_month[fdakma0der_shift$memberships==3,2])
```

using hclust:

```{r}
system.time(
  hclustres <- fdahclust(
  x = alligned_x,
  y = alligned_y,
  n_clusters = 3L,
  warping_class = "none",
  centroid_type = "mean",
  metric = "pearson",
  linkage_criterion = "complete",
  cluster_on_phase = FALSE,
  use_verbose = TRUE,
  warping_options = c(0.15, 0.15),
  maximum_number_of_iterations = 100L,
  number_of_threads = 12L,
  parallel_method = 0L,
  distance_relative_tolerance = 0.001,
  use_fence = FALSE,
  check_total_dissimilarity = TRUE,
  compute_overall_center = FALSE
)
)
```

```{r}
autoplot(hclustres)
```


```{r}
diagnostic_plot(hclustres)
```

this is still not great:

the dominating behaviour is at the end of the month:

```{r}
table(hclustres$memberships)
```


```{r}
matplot(t(hclustres$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:4,lwd = 3)
```

clsuter 1 winter an may

```{r}
table(df_month[hclustres$memberships==1,2])
```

cluster 2 ferb

```{r}
table(df_month[hclustres$memberships==2,2])
```

cluster 3 summer and spring :
```{r}
table(df_month[hclustres$memberships==3,2])
```

trying to solve the february problem by removing the last days an see the effect of
the general  pattern

```{r}
alligning_shift <- fdakmeans(x = x[,1:28],y = y[,1:28], n_clusters = 1,
  seeds = 1,
  warping_class = "shift",
  metric = "pearson",
  centroid_type = "medoid",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F
)

x_all_cut <- alligning_shift$grids[1,]
y_all_cut <- alligning_shift$aligned_curves[,1,]
```

```{r}
k <- 3
system.time(
fdakma0der_shift <- fdakmeans(x = x_all_cut,y = y_all_cut, n_clusters = k,
  seeding_strategy = "kmeans++",
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 10L
)
)
```

```{r}
autoplot(fdakma0der_shift,type = "amplitude")
```

```{r}
matplot(t(fdakma0der_shift$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:3,lwd = 3)
```

cluster 1:

the rest
```{r}
table(df_month[fdakma0der_shift$memberships==1,2])
```

cluster 2:

jaunary, june an september
```{r}
table(df_month[fdakma0der_shift$memberships==2,2])
```

cluster 3:

this is december
```{r}
table(df_month[fdakma0der_shift$memberships==3,2])
```


trying to select k

```{r}
n_sub <- 40
sub_id <- sample(1:n,n_sub,replace = FALSE)
x_sub <- t(matrix(rep(x_all_cut,n_sub),28,n_sub))
y_sub <- as.matrix(y_all_cut[sub_id,])

system.time(invisible(comparison_kmeans <- compare_caps(
  x_sub,
  y_sub,
  n_clusters = 3:6,
  metric = "pearson",
  clustering_method = "kmeans",
  warping_class = "none",
  centroid_type = "mean",
  cluster_on_phase = FALSE
    )
  )
)
```

```{r}
plot(comparison_kmeans, validation_criterion = "wss", what = "mean",lwd = 3)
```

```{r}
plot(comparison_kmeans, validation_criterion = "wss", what = "distribution")
```

```{r}
plot(comparison_kmeans, validation_criterion = "silhouette", what = "mean")
```

```{r}
plot(comparison_kmeans, validation_criterion = "silhouette", what = "distribution")
```

this is suggesting 3 clusters:

using hclust:

```{r}
system.time(
  hclustres <- fdahclust(
  x = x_all_cut,
  y = y_all_cut,
  n_clusters = 3L,
  warping_class = "none",
  centroid_type = "mean",
  metric = "pearson",
  linkage_criterion = "complete",
  cluster_on_phase = FALSE,
  use_verbose = TRUE,
  warping_options = c(0.15, 0.15),
  maximum_number_of_iterations = 100L,
  number_of_threads = 12L,
  parallel_method = 0L,
  distance_relative_tolerance = 0.001,
  use_fence = FALSE,
  check_total_dissimilarity = TRUE,
  compute_overall_center = FALSE
)
)
```

```{r}
autoplot(hclustres)
```

```{r}
diagnostic_plot(hclustres)
```

this is better

the dominating behaviour is at the end of the month:

```{r}
table(hclustres$memberships)
```

```{r}
matplot(t(hclustres$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:4,lwd = 3)
```

cluster 1:

january ad april
```{r}
table(df_month[hclustres$memberships==1,2])
```

cluster 2:

rest of the year
```{r}
table(df_month[hclustres$memberships==2,2])
```

cluster 3:

december
```{r}
table(df_month[hclustres$memberships==3,2])
```

this is the best we have 3 clusters