---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "Functional clustering dayly data"
author:
    - "Valeria Iapaolo"
    - "Oswaldo Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
    
output: html_notebook
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center',
    #fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r libraries inclusions, include=FALSE}
library(tidyverse)
library(roahd)
#library(fda)
library(fdacluster)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/functional/df_hour.RData")
```

first we start with outlier detection in the functional case

```{r}
fd_day <- fData(1:24,as_tibble(df_hour[,2:25]))
```

```{r}
plot(fd_day,lwd = 3,xlab = "day",ylab = "dayly number of crashes",main = "hourly crashes in each week")
```

functional bagplot:

```{r}
hour_fbagplot <- fbplot(fd_day, main="Magnitude outliers horly data")
```

the default F is:

```{r}
hour_fbagplot$Fvalue
```

no outliers found

```{r}
df_hour[hour_fbagplot$ID_outliers,1:2]
```

observations with 0 crashes create a numerical problem, we add 1 to everithing

```{r}
df_hour2 <- df_hour + 1
fd_day2 <- fData(1:24,as_tibble(df_hour[,2:25]))
```

```{r}
hour_fbagplot2 <- fbplot(fd_day2, main="Magnitude outliers week data",
                                  adjust = list( N_trials = 20,trial_size = fd_day2$N,
                                                 VERBOSE = TRUE ))
```

this is not working, don't know why.

chosing a smalle F value:

```{r}
hour_fbagplot2 <- fbplot(fd_day, main="Magnitude outliers week data",Fvalue = 0.9)
```

the chosen F value is:

```{r}
hour_fbagplot2$Fvalue
```

the outlying years are:

```{r}
df_hour[hour_fbagplot2$ID_outliers,1]
```

new yeas eve and 2 strange days : 2005-07-31	and 2009-02-11

outiliergram:

```{r}
invisible(out_hour<- outliergram(fd_day,adjust = F,lwd = 3,display = F))
```

the found outliers are:

```{r}
df_hour[out_hour$ID_outliers,1]
```

the plot of the original function is not working.

adjusting the F:

```{r}
out_hour <- outliergram(fd_day,lwd = 3,adjust = list( N_trials = 20,trial_size = 8*fd_day$N,
                                                 VERBOSE = TRUE ),display = FALSE)
```

```{r}
out_hour$Fvalue
```

nothing changed, same outliers detected.

```{r}
df_hour[out_hour$ID_outliers,1:2]
```

plotting in the old way.

```{r}
par(mfrow=c(1,2))
plot(fd_day[out_hour$ID_outliers,],lwd = 1,main = "outliers",col = 2)
plot(fd_day[-out_hour$ID_outliers,],lwd = 1,main = "non outliers",col = 3)
```

there are 2 typer of outtliers:
very low crashes days,
very high crashes day.

let's do some clustering:

```{r}
hours <- 1:24
n <- fd_day$N
x <- t(matrix(rep(hours,n),24,n))
y <- as.matrix(df_hour[,2:25])
```

```{r}
k <- 3
system.time(
fdakma0der <- fdakmeans(x = x,y = y, n_clusters = k,
  seeds = sample(1:n,k),
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 12
)
)
```

```{r}
autoplot(fdakma0der)
```

```{r}
matplot(t(fdakma0der$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:5,lwd = 3)
```

```{r}
k <- 3
system.time(
fdakma0der <- fdakmeans(x = x,y = y, n_clusters = k,
  seeds = sample(1:n,k),
  warping_class = "none",
  metric = "l2",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 12
)
)
```

```{r}
autoplot(fdakma0der)
```

```{r}
matplot(t(fdakma0der$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:5,lwd = 3)
```

pearson is the prefered metric, since it takes into account the shape as well.

we need to chose the k:

```{r}
n_sub <- 50
sub_id <- sample(1:n,n_sub,replace = FALSE)
x_sub <- x[sub_id,]
y_sub <- y[sub_id,]

system.time(invisible(comparison_kmeans <- compare_caps(
  x_sub,
  y_sub,
  n_clusters = 1:4,
  metric = "pearson",
  clustering_method = "kmeans",
  warping_class = "none",
  centroid_type = "mean",
  cluster_on_phase = FALSE
    )
  )
)
```

```{r}
plot(comparison_kmeans, validation_criterion = "wss", what = "mean",lwd = 3)
```

```{r}
plot(comparison_kmeans, validation_criterion = "wss", what = "distribution")
```

```{r}
plot(comparison_kmeans, validation_criterion = "silhouette", what = "mean")
```

```{r}
plot(comparison_kmeans, validation_criterion = "silhouette", what = "distribution")
```

it is not clear, k = 2

```{r}
k <- 2
system.time(
fdakma0der <- fdakmeans(x = x,y = y, n_clusters = k,
  seeds = sample(1:n,k),
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 20L
)
)
```

```{r}
autoplot(fdakma0der)
```

```{r}
matplot(t(fdakma0der$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:5,lwd = 3)
```

this is a clear distinction between working days and holydays.

trying k = 4

```{r}
k <- 4
system.time(
fdakma0der <- fdakmeans(x = x,y = y, n_clusters = k,
  seeds = sample(1:n,k),
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 20L
)
)
```

```{r}
autoplot(fdakma0der)
```

```{r}
matplot(t(fdakma0der$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:5,lwd = 3)
```

it is clear that this is just a magnitude thing for the first peaks

trying k = 10

```{r}
k <- 10
system.time(
fdakma0der <- fdakmeans(x = x,y = y, n_clusters = k,
  seeds = sample(1:n,k),
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 20L
)
)
```

```{r}
autoplot(fdakma0der)
```

```{r}
matplot(t(fdakma0der$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:k,lwd = 3)
```

not much more information here, stick to 2




