---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "Functional clustering"
author:
    - "Valeria Iapaolo"
    - "Oswaldo Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
output:
    pdf_document:
        toc: true
        toc_depth: 3
        number_section: true
    html_document:
        toc: true
        toc_float: true
        number_sections: true
date: "2023-11-10"
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    dev = c('pdf'),
    fig.align = 'center',
    fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r libraries inclusions, include=FALSE}
library(tidyverse)
library(roahd)
library(fdacluster)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/functional/df_year.RData")
```

first we start with outlier detection in the functional case

```{r}
fd_year <- fData(1:366,df_year[,2:367])
```

```{r}
plot(fd_year,lwd = 3,xlab = "day",ylab = "number of crashes per day",main = "dayly crashes in each year")
```

functional bagplot:

```{r}
year_fbagplot <- fbplot(fd_year, main="Magnitude outliers yearly data")
```

the default F is:

```{r}
year_fbagplot$Fvalue
```

the outlying years are:

```{r}
df_year[year_fbagplot$ID_outliers,1]
```

we can see that the covi year is an outliers as are the firs year where the number of total crashes is much higher than in the next years.

if we try to adjust the inflation factor:

we need to remove the day 366 since it causes problems in the iteerations to find the F:

```{r}
fd_year2 <- fData(1:365,df_year[,2:366])
```

```{r}
year_fbagplot2 <- fbplot(fd_year2, main="Magnitude outliers yearly data",
                                  adjust = list( N_trials = 20,trial_size = 50,
                                                 VERBOSE = TRUE ))
```

the chosen F value is:

```{r}
year_fbagplot2$Fvalue
```

the outlying years are:

```{r}
df_year[year_fbagplot2$ID_outliers,1]
```

nothing changed.

outiliergram:

```{r}
invisible(out_year <- outliergram(fd_year,adjust = F,lwd = 3))
```

no outliers found, adjusting the F:

```{r}
out_year2 <- outliergram(fd_year2,lwd = 5,adjust = list( N_trials = 20,trial_size = 50,
                                                 VERBOSE = TRUE ))
```

```{r}
out_year2$Fvalue
```

nothing changed, no outliers detected.

we could can try to do clustering of the functional data:

```{r}
n <- 18
days <- 1:366
x <- t(matrix(rep(days,n),365,n))
y <- as.matrix(df_year[,2:366])
k <- 2

fdakma0der_shift <- fdakmeans(x = x,y = y, n_clusters = k,
  seeding_strategy = "kmeans++",
  warping_class = "shift",
  metric = "l2",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = TRUE
)
```

```{r}
autoplot(fdakma0der_shift,type = "amplitude")
```

```{r}
autoplot(fdakma0der_shift,type = "phase")
```

```{r}
diagnostic_plot(fdakma0der_shift)
```

this is not great,

trying with the affine warping:

```{r}
fdakma0der_affine <- fdakmeans(x = x,y = y, n_clusters = k,
  seeding_strategy = "kmeans++",
  warping_class = "affine",
  metric = "l2",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = TRUE
)
```

```{r}
autoplot(fdakma0der_affine,type = "amplitude")
```

```{r}
autoplot(fdakma0der_affine,type = "phase")
```


```{r}
diagnostic_plot(fdakma0der_affine)
```

using perason instead of l2 distance:

```{r}
fdakma0der_shift <- fdakmeans(x = x,y = y, n_clusters = k,
  seeding_strategy = "kmeans++",
  warping_class = "shift",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = TRUE
)
```

```{r}
autoplot(fdakma0der_shift,type = "amplitude")
```

```{r}
autoplot(fdakma0der_shift,type = "phase")
```

this gets the single outlier, the other distance did not because of the oscillations.

```{r}
fdakma0der_shift <- fdakmeans(x = x,y = y, n_clusters = 4,
  seeding_strategy = "kmeans++",
  warping_class = "shift",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = TRUE
)
```

```{r}
autoplot(fdakma0der_shift)
```

maybe I should cluster the derivatives for a better pattern, but needs smoothing for that.

```{r}
df_year[which(fdakma0der_shift$memberships==1),1]
```

```{r}
df_year[which(fdakma0der_shift$memberships==2),1]
```

```{r}
df_year[which(fdakma0der_shift$memberships==3),1]
```

```{r}
df_year[which(fdakma0der_shift$memberships==4),1]
```


we canot se a clear clustering in the data.