---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "Functional clustering monthly data"
author:
    - "Valeria Iapaolo"
    - "Oswaldo Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
    
output: html_notebook
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center',
    #fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r libraries inclusions, include=FALSE}
library(tidyverse)
library(roahd)
#library(fda)
library(fdacluster)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/functional/df_week.RData")
```

first we start with outlier detection in the functional case

```{r}
fd_week <- fData(1:7,as_tibble(df_week[,3:9]))
```

```{r}
plot(fd_week,lwd = 3,xlab = "day",ylab = "dayly number of crashes",main = "dayly crashes in each week")
```

functional bagplot:

```{r}
week_fbagplot <- fbplot(fd_week, main="Magnitude outliers weekly data")
```

the default F is:

```{r}
week_fbagplot$Fvalue
```


no outliers found

```{r}
df_week[week_fbagplot$ID_outliers,1:2]
```


we need to remove the weeks that cause problems witth 0 (1,52,53)

```{r}
df_week2 <- df_week %>% filter(week_of_year %in% 2:51)
fd_week2 <- fData(1:7,as_tibble(df_week2[,3:9]))
```

```{r}
week_fbagplot2 <- fbplot(fd_week2, main="Magnitude outliers week data",
                                  adjust = list( N_trials = 20,trial_size = fd_week2$N,
                                                 VERBOSE = TRUE ))
```

the chosen F value is:

```{r}
week_fbagplot2$Fvalue
```

the outlying years are:

```{r}
df_week2[week_fbagplot2$ID_outliers,1:2]
```

first years ando covid year

outiliergram:

```{r}
invisible(out_week<- outliergram(fd_week,adjust = F,lwd = 3,display = F))
```

the found outliers are:

```{r}
df_week[out_week$ID_outliers,1:2]
```

the plot of the original function is not working.

adjusting the F:

```{r}
out_week <- outliergram(fd_week,lwd = 3,adjust = list( N_trials = 20,trial_size = 8*fd_week$N,
                                                 VERBOSE = TRUE ),display = FALSE)
```

```{r}
out_week$Fvalue
```

nothing changed, same outliers detected.

```{r}
df_week[out_week$ID_outliers,1:2]
```

plotting in the old way.

```{r}
par(mfrow=c(1,2))
plot(fd_week[out_week$ID_outliers,],lwd = 1,main = "outliers",col = 2)
plot(fd_week[-out_week$ID_outliers,],lwd = 1,main = "non outliers",col = 3)
```

trying to remove the weeks with the zeros:

```{r}
week2_fbagplot <- fbplot(fd_week2, main="Magnitude outliers weekly data")
```

the default F is:

```{r}
week2_fbagplot$Fvalue
```

no outliers found

```{r}
df_week[week2_fbagplot$ID_outliers,1:2]
```

```{r}
week_fbagplot2 <- fbplot(fd_week2, main="Magnitude outliers week data",
                                  adjust = list( N_trials = 20,trial_size = fd_week2$N,
                                                 VERBOSE = TRUE ))
```

the chosen F value is:

```{r}
week_fbagplot2$Fvalue
```

the outlying years are:

```{r}
df_week2[week_fbagplot2$ID_outliers,1:2]
```

first years ando covid year

outiliergram:

```{r}
invisible(out_week2<- outliergram(fd_week2,adjust = F,lwd = 3,display = F))
```

the found outliers are:

```{r}
df_week[out_week$ID_outliers,1:2]
```

the plot of the original function is not working.

adjusting the F:

```{r}
out_week2 <- outliergram(fd_week2,lwd = 3,adjust = list( N_trials = 20,trial_size = 8*fd_week$N,
                                                 VERBOSE = TRUE ),display = FALSE)
```

```{r}
out_week2$Fvalue
```

nothing changed, same outliers detected.

```{r}
df_week[out_week2$ID_outliers,1:2]
```

plotting in the old way.

```{r}
par(mfrow=c(1,2))
plot(fd_week[out_week2$ID_outliers,],lwd = 1,main = "outliers",col = 2)
plot(fd_week[-out_week2$ID_outliers,],lwd = 1,main = "non outliers",col = 3)
```

we can do some clustering, no need to allign the data this time:

```{r}
days <- 1:7
n <- fd_week$N
x <- t(matrix(rep(days,n),7,n))
y <- as.matrix(df_week[,3:9])
```

tryig k = 3

```{r}
k <- 3
system.time(
fdakma0der <- fdakmeans(x = x,y = y, n_clusters = k,
  seeds = sample(1:n,k),
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 12
)
)
```

```{r}
autoplot(fdakma0der,type = "amplitude")
```

selecting the number of clusters:

```{r}
n_sub <- 50
sub_id <- sample(1:n,n_sub,replace = FALSE)
x_sub <- x[sub_id,]
y_sub <- y[sub_id,]

system.time(invisible(comparison_kmeans <- compare_caps(
  x_sub,
  y_sub,
  n_clusters = 2:5,
  metric = "pearson",
  clustering_method = "kmeans",
  warping_class = "none",
  centroid_type = "mean",
  cluster_on_phase = FALSE
    )
  )
)
```

```{r}
plot(comparison_kmeans, validation_criterion = "wss", what = "mean",lwd = 3)
```

```{r}
plot(comparison_kmeans, validation_criterion = "wss", what = "distribution")
```

```{r}
plot(comparison_kmeans, validation_criterion = "silhouette", what = "mean")
```

```{r}
plot(comparison_kmeans, validation_criterion = "silhouette", what = "distribution")
```


2 is probably the best, lets take a look:

```{r}
k <- 2
system.time(
fdakma0der <- fdakmeans(x = x,y = y, n_clusters = k,
  seeds = sample(1:n,k),
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 12
)
)
```

```{r}
autoplot(fdakma0der,type = "amplitude")
```

```{r}
table(df_week[fdakma0der$memberships==1,2])
```

```{r}
table(df_week[fdakma0der$memberships==2,2])
```

```{r}
id1 <- which(fdakma0der$memberships==1)
id2 <- which(fdakma0der$memberships==2)
par(mfrow=c(1,2))
plot(fd_week[id1[1:30],],lwd = 1,main = "cluster 1",col = 2)
plot(fd_week[id2[1:30],],lwd = 1,main = "cluster 2",col = 3)
```

this is probably holiday vs not holiday

```{r}
n2 <- fd_week2$N
x2 <- t(matrix(rep(days,n2),7,n2))
y2 <- as.matrix(df_week2[,3:9])
```

```{r}
k <- 2
system.time(
fdakma0der <- fdakmeans(x = x2,y = y2, n_clusters = k,
  seeds = sample(1:n,k),
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 12
)
)
```

```{r}
autoplot(fdakma0der,type = "amplitude")
```

```{r}
table(df_week2[fdakma0der$memberships==1,2])
```

```{r}
table(df_week2[fdakma0der$memberships==2,2])
```

```{r}
id1 <- which(fdakma0der$memberships==1)
id2 <- which(fdakma0der$memberships==2)
par(mfrow=c(1,2))
plot(fd_week2[id1[1:30],],lwd = 1,main = "cluster 1",col = 2)
plot(fd_week2[id2[1:30],],lwd = 1,main = "cluster 2",col = 3)
```

no difference for excluding the problematic weeks.

```{r}
k <- 3
system.time(
fdakma0der <- fdakmeans(x = x2,y = y2, n_clusters = k,
  seeds = sample(1:n,k),
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 12
)
)
```

```{r}
autoplot(fdakma0der,type = "amplitude")
```

```{r}
table(df_week2[fdakma0der$memberships==1,2])
```

```{r}
table(df_week2[fdakma0der$memberships==2,2])
```

```{r}
table(df_week2[fdakma0der$memberships==3,2])
```


```{r}
id1 <- which(fdakma0der$memberships==1)
id2 <- which(fdakma0der$memberships==2)
id3 <- which(fdakma0der$memberships==3)
par(mfrow=c(1,3))
plot(fd_week2[id1[1:30],],lwd = 1,main = "cluster 1",col = 2)
plot(fd_week2[id2[1:30],],lwd = 1,main = "cluster 2",col = 3)
plot(fd_week2[id3[1:30],],lwd = 1,main = "cluster 3",col = 4)
```

2 an 3 are very similar,

trying with 4

```{r}
k <- 4
system.time(
fdakma0der <- fdakmeans(x = x2,y = y2, n_clusters = k,
  seeds = sample(1:n,k),
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 12
)
)
```

```{r}
autoplot(fdakma0der,type = "amplitude")
```

```{r}
table(df_week2[fdakma0der$memberships==1,2])
```

```{r}
table(df_week2[fdakma0der$memberships==2,2])
```

```{r}
table(df_week2[fdakma0der$memberships==3,2])
```

```{r}
table(df_week2[fdakma0der$memberships==4,2])
```

```{r}
id1 <- which(fdakma0der$memberships==1)
id2 <- which(fdakma0der$memberships==2)
id3 <- which(fdakma0der$memberships==3)
id4 <- which(fdakma0der$memberships==4)
par(mfrow=c(2,2))
plot(fd_week2[id1[1:30],],lwd = 1,main = "cluster 1",col = 2)
plot(fd_week2[id2[1:30],],lwd = 1,main = "cluster 2",col = 3)
plot(fd_week2[id3[1:30],],lwd = 1,main = "cluster 3",col = 4)
plot(fd_week2[id4[1:30],],lwd = 1,main = "cluster 4",col = 5)
```


still better with 2.

trying hierarchical clustering

```{r}
system.time(
  hclustres <- fdahclust(
  x = x2,
  y = y2,
  n_clusters = 2L,
  warping_class = "none",
  centroid_type = "mean",
  metric = "pearson",
  linkage_criterion = "complete",
  cluster_on_phase = FALSE,
  use_verbose = TRUE,
  warping_options = c(0.15, 0.15),
  maximum_number_of_iterations = 100L,
  number_of_threads = 12L,
  parallel_method = 0L,
  distance_relative_tolerance = 0.001,
  use_fence = FALSE,
  check_total_dissimilarity = TRUE,
  compute_overall_center = FALSE
)
)
```

```{r}
autoplot(hclustres)
```

```{r}
diagnostic_plot(hclustres)
```

```{r}
table(hclustres$memberships)
```

```{r}
matplot(t(hclustres$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:5,lwd = 3)
```

the pattern is always the same, division of the two clusters.

k-menas on the whole dataset:

```{r}
k <- 2
system.time(
fdakma0der <- fdakmeans(x = x,y = y, n_clusters = k,
  seeds = sample(1:n,k),
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 12
)
)
```

```{r}
autoplot(fdakma0der,type = "amplitude")
```

```{r}
table(df_week[fdakma0der$memberships==1,2])
```

```{r}
table(df_week[fdakma0der$memberships==2,2])
```

```{r}
id1 <- which(fdakma0der$memberships==1)
id2 <- which(fdakma0der$memberships==2)
par(mfrow=c(1,2))
plot(fd_week[id1[1:30],],lwd = 1,main = "cluster 1",col = 2)
plot(fd_week[id2[1:30],],lwd = 1,main = "cluster 2",col = 3)
```

plotting the weeks differently:

```{r}
df <- data.frame(table(df_week[fdakma0der$memberships==1,2]),table(df_week[fdakma0der$memberships==2,2])) %>% select(-week_of_year.1)
df %>% ggplot(aes(order(week_of_year),Freq)) + geom_line()
```

this shows that group 1 are the work week, while cluster 2 are the holiays.