---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "Functional permutation tests monthly data"
author:
    - "Valeria Iapaolo"
    - "Oswaldo Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
    
output: html_notebook
---


```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center',
    #fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r libraries inclusions, include=FALSE}
library(tidyverse)
library(roahd)
#library(fda)
library(fdacluster)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/functional/df_month.RData")
```

first we start with outlier detection in the functional case

```{r}
fd_month <- fData(1:31,df_month[,3:33])
```

```{r}
plot(fd_month,lwd = 3,xlab = "day",ylab = "dayly number of crashes",main = "dayly crashes in each month")
```

functional bagplot:

```{r}
year_fbagplot <- fbplot(fd_month, main="Magnitude outliers monthly data")
```

the default F is:

```{r}
year_fbagplot$Fvalue
```

no outliers found

```{r}
df_month[year_fbagplot$ID_outliers,1:2]
```

we need to remove the day 366 since it causes problems in the iterations to find the F:

```{r}
fd_month2 <- fData(1:28,df_month[,3:30])
```

```{r}
year_fbagplot2 <- fbplot(fd_month, main="Magnitude outliers yearly data",
                                  adjust = list( N_trials = 20,trial_size = fd_month2$N,
                                                 VERBOSE = TRUE ))
```

the chosen F value is:

```{r}
year_fbagplot2$Fvalue
```

the outlying years are:

```{r}
df_month[year_fbagplot2$ID_outliers,1:2]
```

nothing changedjust the first year are seen as outliers in this case

outiliergram:

```{r}
invisible(out_month <- outliergram(fd_month,adjust = F,lwd = 3,display = T))
```

the found outliers are:

```{r}
df_month[out_month$ID_outliers,1:2]
```

this are all in december.

the plot of the original function is not working.

adjusting the F:

```{r}
out_month <- outliergram(fd_month,lwd = 5,adjust = list( N_trials = 20,trial_size = 8*fd_month$N,
                                                 VERBOSE = TRUE ),display = FALSE)
```

```{r}
out_month$Fvalue
```

nothing changed, same outliers detected.

```{r}
df_month[out_month$ID_outliers,1:2]
```

plotting in the old way.

```{r}
par(mfrow=c(1,2))
plot(fd_month[out_month$ID_outliers,],lwd = 3,main = "outliers",col = 2)
plot(fd_month[-out_month$ID_outliers,],lwd = 3,main = "non outliers",col = 3)
```

there is a clear pattern in the data due to the christmas period.

doing some clustering:

let's run a comparison of the methods:

```{r}
days <- 1:31
n <- fd_month$N
x <- t(matrix(rep(days,n),31,n))
y <- as.matrix(df_month[,3:33])
```

another possibility is to align all of the curves before the clustering and then do the clustering:

```{r}
fdakma0der_srsf <- fdakmeans(x = x,y = y, n_clusters = 1,
  seeds = 1,
  warping_class = "srsf",
  metric = "pearson",
  centroid_type = "medoid",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F
)

```

```{r}
autoplot(fdakma0der_srsf,type = "amplitude")
```

```{r}
autoplot(fdakma0der_srsf,type = "phase")
```

we recover the curves in the following way:

```{r}
alligned_x <- fdakma0der_srsf$grids[1,]
alligned_y <- fdakma0der_srsf$aligned_curves[,1,]
```

```{r}
plot(fData(alligned_x,alligned_y))
abline(v = c(3,10,17,24))
```

```{r}
fdakma0der_shift <- fdakmeans(x = x,y = y, n_clusters = 1,
  seeds = 1,
  warping_class = "shift",
  metric = "pearson",
  centroid_type = "medoid",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  maximum_number_of_iterations = 1
)

```

```{r}
autoplot(fdakma0der_shift,type = "amplitude")
```

```{r}
autoplot(fdakma0der_shift,type = "phase")
```

we recover the curves in the following way:

```{r}
alligned_x <- fdakma0der_shift$grids[1,]
alligned_y <- fdakma0der_shift$aligned_curves[,1,]
```

```{r}
plot(fData(alligned_x,alligned_y))
```

```{r}
days2 <- 1:28
n <- fd_month$N
x2 <- t(matrix(rep(days2,n),28,n))
y2 <- as.matrix(df_month[,3:30])
```

another possibility is to align all of the curves before the clustering and then do the clustering:

```{r}
fdakma0der_shift2 <- fdakmeans(x = x2,y = y2, n_clusters = 1,
  seeds = 1,
  warping_class = "shift",
  metric = "pearson",
  centroid_type = "medoid",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  maximum_number_of_iterations = 1
)

```

```{r}
autoplot(fdakma0der_shift2,type = "amplitude")
```

```{r}
autoplot(fdakma0der_shift2,type = "phase")
```

we recover the curves in the following way:

```{r}
alligned_x2 <- fdakma0der_shift2$grids[1,]
alligned_y2 <- fdakma0der_shift2$aligned_curves[,1,]
```

```{r}
plot(fData(alligned_x2,alligned_y2))
```

this clearly works better

we can cluster this:

```{r}
k <- 3
system.time(
fdakma0der_shift <- fdakmeans(x = alligned_x2,y = alligned_y2, n_clusters = k,
  seeding_strategy = "kmeans++",
  warping_class = "none",
  metric = "pearson",
  centroid_type = "mean",
  distance_relative_tolerance = 1e-3,
  add_silhouettes = F,
  parallel_method = 0L,
  number_of_threads = 10L
)
)
```

```{r}
autoplot(fdakma0der_shift,type = "amplitude")
```

```{r}
i1 <- which(fdakma0der_shift$memberships==1)
i2 <- which(fdakma0der_shift$memberships==2)
i3 <- which(fdakma0der_shift$memberships==3)


n <- dim(df_month)[1]
indexes_plot <- c(i1[sample(length(i1),60)],i2[sample(length(i2),30)],i3)
nplots <- length(indexes_plot)
unicurves <- fdakma0der_shift$original_curves[indexes_plot, 1, ]
grids <- matrix(rep(1:28,nplots),nrow = nplots, ncol = 28,byrow = TRUE)
memberships <- fdakma0der_shift$memberships[indexes_plot]

wrangled_data <- tibble::tibble(
      grid = purrr::array_tree(grids, margin = 1),
      value = purrr::array_tree(unicurves, margin = 1),
      membership = memberships,
      curve_id = 1:nplots
    ) |>
      tidyr::unnest(cols = c(.data$grid, .data$value)) |> 
  dplyr::bind_rows() |>
    dplyr::mutate(
      membership = as.factor(.data$membership),
      curve_id = as.factor(.data$curve_id),
    )

```

```{r}
P1 <- wrangled_data |>
      ggplot2::ggplot(ggplot2::aes(
        x = .data$grid,
        y = .data$value,
        color = .data$membership,
        group = .data$curve_id
      )) +
      ggplot2::geom_line() +
      ggplot2::facet_wrap(ggplot2::vars(.data$membership)
      ) +
      ggplot2::labs(
        x = "Days",
        y = "number of collisions per day"
      ) +
      ggplot2::theme(legend.position = "none")

P1

```


```{r}
matplot(t(fdakma0der_shift$center_curves[,1,]),type = 'l',
        main='clustered and alligned curves',xlab='days',ylab='crashes', col = 1:3,lwd = 3)
```

cluster 1:

the rest
```{r}
table(df_month[fdakma0der_shift$memberships==1,2])
```

cluster 2:

jaunary mostly
```{r}
table(df_month[fdakma0der_shift$memberships==2,2])
```

cluster 3:

this is december
```{r}
table(df_month[fdakma0der_shift$memberships==3,2])
```

we can now so some permutation tests:

```{r}
library(fdatest)
```

I start with a permutation test on the global curves:

```{r}
i1 <- which(fdakma0der_shift$memberships==1)
n1 <- length(i1)

i2 <- which(fdakma0der_shift$memberships==2)
n2 <- length(i2)
i3 <- which(fdakma0der_shift$memberships==3)
n3 <- length(i3)
```

```{r}
seed=7122023
B=1000
```

start by comparing group 1 and 2:

```{r}
aug_df=rbind(alligned_y2[i1,],alligned_y2[i2,])
n= n1 + n2 

meandiff=(colMeans(alligned_y2[i1,])-colMeans(alligned_y2[i2,]))
plot(meandiff,type = 'l')
T0=sum(meandiff^2)
T0
```

And, Knowing that under $H_0$ the two groups of curves are IID, my likelihood-invariant permutation scheme is of course label permutation, so...

```{r}
T0_perm=numeric(B)

for(perm in 1:B){
  permutation <- sample(n)
  df_perm=aug_df[permutation,]
  perm_1 = df_perm[1:n1,] 
  perm_2 = df_perm[(n1+1):n,] 
  T0_perm[perm]=sum(((colMeans(perm_1)-colMeans(perm_2)))^2)
}

sum(T0_perm >= T0)/B

hist(T0_perm,xlim = c(0,2*T0))
abline(v=T0,col='green')
```

they are significantly different

doing the same for cluster 1 an 3.

```{r}
aug_df=rbind(alligned_y2[i1,],alligned_y2[i3,])
n= n1 + n3 

meandiff=(colMeans(alligned_y2[i1,])-colMeans(alligned_y2[i3,]))
plot(meandiff,type = 'l')
T0=sum(meandiff^2)
T0
```

```{r}
T0_perm=numeric(B)

for(perm in 1:B){
  permutation <- sample(n)
  df_perm=aug_df[permutation,]
  perm_1 = df_perm[1:n1,] 
  perm_2 = df_perm[(n1+1):n,] 
  T0_perm[perm]=sum(((colMeans(perm_1)-colMeans(perm_2)))^2)
}

sum(T0_perm >= T0)/B

hist(T0_perm,xlim = c(0,2*T0))
abline(v=T0,col='green')
```

statistically different also.

clustr 2 an 3:

```{r}
aug_df=rbind(alligned_y2[i2,],alligned_y2[i3,])
n= n2 + n3 

meandiff=(colMeans(alligned_y2[i2,])-colMeans(alligned_y2[i3,]))
plot(meandiff,type = 'l')
T0=sum(meandiff^2)
T0
```

And, Knowing that under $H_0$ the two groups of curves are IID, my likelihood-invariant permutation scheme is of course label permutation, so...

```{r}
T0_perm=numeric(B)

for(perm in 1:B){
  permutation <- sample(n)
  df_perm=aug_df[permutation,]
  perm_1 = df_perm[1:n2,] 
  perm_2 = df_perm[(n2+1):n,] 
  T0_perm[perm]=sum(((colMeans(perm_1)-colMeans(perm_2)))^2)
}

sum(T0_perm >= T0)/B

hist(T0_perm,xlim = c(0,2*T0))
abline(v=T0,col='green')
```

same here.

we now move to a local test.

we use the interval wise procedure, so we control the interval-wise error rate

between clusters 1 an 2

```{r}
tst=IWT2(alligned_y2[i1,],alligned_y2[i2,])
plot(tst)
```

we can see that the ifference is significant at the beginnig of the interval,this makes sence cluster 1
is the general yesr wile cluster 2 is january.

doing the same for cluster 1 an 3:

```{r}
tst=IWT2(alligned_y2[i1,],alligned_y2[i3,])
plot(tst)
```

the difference is just at the end because of the christmas breake, which is at the end of the month of december which is cluster 3

and for cluster 2 and 3 

```{r}
tst=IWT2(alligned_y2[i2,],alligned_y2[i3,])
plot(tst)
```

as expected the difference is at the eges of the domains. 

we could o functional analysys of variance:

```{r}
groups <- fdakma0der_shift$memberships

faov <- ITPaovbspline(alligned_y2 ~ groups)
```

```{r}
summary(faov)
```

```{r}
plot(faov)
```

as expected the difference is in the tails as well as the head
