---
title: "Nonparametric Analysis of UK Road Accidents"
subtitle: "functional conformal prediction for each police district"
author:
    - "Valeria Iapaolo"
    - "Oswaldo Morales"
    - "Riccardo Morandi"
    - "Abylai Orynbassar"
output: html_notebook
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    #dev = c('pdf'),
    fig.align = 'center',
    #fig.path = 'output/',
    fig.height = 6,
    fig.width = 12
)
```

```{r}
library(tidyverse)
library(conformalInference.fd)
library(roahd)
library(lubridate)
library(mgcv)
library(mgcViz)
```

```{r}
load("~/Documents/Nonparametric Statisics/Project/clean data/full_collisions.RData")
```

```{r}
glimpse(full_collisions)
```

we need to create the dataset in a way that they cotain all of the days and a zero if no crashes happened in that day:

```{r}
all_dates <- unique(full_collisions$date)
all_police_forces <- unique(full_collisions$police_force)

full_combinations <- expand.grid(date = all_dates, police_force = all_police_forces)

crashes_per_day_police <- full_combinations %>%
  left_join(full_collisions %>% group_by(date, police_force) %>% summarize(number_of_crashes = n()),
            by = c("date", "police_force")) %>%
  replace_na(list(number_of_crashes = 0))

glimpse(crashes_per_day_police)
```

```{r}
crashes_per_day_police <- crashes_per_day_police %>% mutate(day_of_year = yday(date),
                                                            day = factor(wday(date)),year = year(date))
```

removing the 29 of february for dimensioality problems in the vectors below

```{r}
crashes_per_day_police <- crashes_per_day_police %>% filter(!(month(date)==2 & day(date)==29))
```

```{r}
train_years <- ceiling(seq(2005,2021,by = 1.5))
cal_years <- seq(2006,2021,3)

df_train <- crashes_per_day_police %>% filter(year %in% train_years) 
df_cal <- crashes_per_day_police %>% filter(year %in% cal_years) 
df_test <- crashes_per_day_police %>% filter(year == 2022)
```

fitting a gam for each police force:

```{r}
gams <- list()

S = matrix(nrow = n_pol,ncol = 365)
k = rep(0,n_pol)

for(i in 1:n_pol){
  
  df_train_mid <- df_train %>% filter(police_force == levels(police_force)[i])
  df_cal_mid <- df_cal %>% filter(police_force == levels(police_force)[i])
  
  gam <- bam(number_of_crashes ~ day + year +
                                  s(day_of_year,k = 53, bs = "cr"), 
                data = df_train_mid, family=quasipoisson(), method='REML')
  
  res_train_mid <- cbind(df_train_mid,preds = predict(gam, df_train_mid, type = "response")) %>%
  arrange(date) %>% mutate(absdiff = abs(number_of_crashes-preds))

  res_cal_mid <- cbind(df_cal_mid,preds = predict(gam, df_cal_mid, type = "response")) %>%
  arrange(date) %>% mutate(absdiff = abs(number_of_crashes-preds))
  
  
  ad_train <- res_train_mid %>% filter(police_force == levels(police_force)[i]) %>%
  dplyr::select(absdiff)
  a_mat <- matrix(ad_train$absdiff,byrow = F,nrow = 365)
  ncm = apply(a_mat,1,max)
  den = line_integral(x, ncm)
  s = ncm/den
  S[i,] = s
  ad_cal <- res_cal_mid %>% filter(police_force == levels(police_force)[i]) %>%
  dplyr::select(absdiff)
  a_mat <- matrix(ad_cal$absdiff,byrow = F,nrow = 365)
  ncs = apply(a_mat/S[i,],1,max)
  k[i] = max(ncs)
  
  gams[[i]] <- gam
}
```

midlands

```{r}
i <- 50
df_test_filtered <- df_test %>% filter(police_force == levels(police_force)[i])
preds <- predict(gams[[i]], df_test_filtered, type = "response")
df_test_filtered  <- cbind(df_test_filtered,preds) %>% arrange(day_of_year)
```

```{r}
df_plot <- data.frame(n = df_test_filtered$number_of_crashes,pred = df_test_filtered$preds, 
                      ymax = df_test_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_test_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])
```

London:

```{r}
i <- 30
df_test_filtered <- df_test %>% filter(police_force == levels(police_force)[i])
preds <- predict(gams[[i]], df_test_filtered, type = "response")
df_test_filtered  <- cbind(df_test_filtered,preds) %>% arrange(day_of_year)
```

```{r}
df_plot <- data.frame(n = df_test_filtered$number_of_crashes,pred = df_test_filtered$preds, 
                      ymax = df_test_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_test_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])
```

the model is still no great

we probably need more data.

using all the available data:

```{r}
full_collisioin_data <- readr::read_csv(file="dft-road-casualty-statistics-collision-1979-latest-published-year.csv") 

spec(full_collisioin_data)
glimpse(full_collisioin_data)

full_collisions_all <- stats19::format_collisions(full_collisioin_data)

glimpse(full_collisions_all)
```

```{r}
all_dates <- unique(full_collisions_all$date)
all_police_forces <- unique(full_collisions_all$police_force)

full_combinations <- expand.grid(date = all_dates, police_force = all_police_forces)

crashes_per_day_police <- full_combinations %>%
  left_join(full_collisions_all %>% group_by(date, police_force) %>% 
              summarize(number_of_crashes = n()),
            by = c("date", "police_force")) %>%
  replace_na(list(number_of_crashes = 0))

glimpse(crashes_per_day_police)
```

```{r}
crashes_per_day_police <- crashes_per_day_police %>% mutate(day_of_year = yday(date),
                                                            day = factor(wday(date)),year = year(date))
```

removing the 29 of february for dimensioality problems in the vectors below

```{r}
crashes_per_day_police <- crashes_per_day_police %>% filter(!(month(date)==2 & day(date)==29))
```

```{r}
train_years <- ceiling(seq(1979,2021,by = 1.5))
cal_years <- seq(1979,2021,3)

df_train <- crashes_per_day_police %>% filter(year %in% train_years) 
df_cal <- crashes_per_day_police %>% filter(year %in% cal_years) 
df_test <- crashes_per_day_police %>% filter(year == 2022)
```

doing it for a single police force at a time:

```{r}
gams <- list()

S = matrix(nrow = n_pol,ncol = 365)
k = rep(0,n_pol)

for(i in 1:n_pol){
  
  df_train_mid <- df_train %>% filter(police_force == levels(police_force)[i])
  df_cal_mid <- df_cal %>% filter(police_force == levels(police_force)[i])
  
  gam <- bam(number_of_crashes ~ day + year +
                                  s(day_of_year,k = 53, bs = "cr"), 
                data = df_train_mid, family=quasipoisson(), method='REML')
  
  res_train_mid <- cbind(df_train_mid,preds = predict(gam, df_train_mid, type = "response")) %>%
  arrange(date) %>% mutate(absdiff = abs(number_of_crashes-preds))

  res_cal_mid <- cbind(df_cal_mid,preds = predict(gam, df_cal_mid, type = "response")) %>%
  arrange(date) %>% mutate(absdiff = abs(number_of_crashes-preds))
  
  ad_train <- res_train_mid %>% filter(police_force == levels(police_force)[i]) %>%
  dplyr::select(absdiff)
  a_mat <- matrix(ad_train$absdiff,byrow = F,nrow = 365)
  setcfr <- apply(a_mat,2,max)
  gamma <- sort(setcfr)[ceiling((m+1)*(1-alpha))]
  a_mat_filt <- a_mat[,setcfr<=gamma]
  ncm = apply(a_mat_filt,1,max)
  ncm_sort <- sort(ncm)
  den = line_integral(x, ncm)
  s = ncm/den
  S[i,] = s
  
  ad_cal <- res_cal_mid %>% filter(police_force == levels(police_force)[i]) %>%
  dplyr::select(absdiff)
  a_mat <- matrix(ad_cal$absdiff,byrow = F,nrow = 365)
  ncs = apply(a_mat/S[i,],1,max)
  k[i] = max(ncs)
  
  gams[[i]] <- gam
}
```

midlands

```{r}
i <- 50
df_test_filtered <- df_test %>% filter(police_force == levels(police_force)[i])
preds <- predict(gams[[i]], df_test_filtered, type = "response")
df_test_filtered  <- cbind(df_test_filtered,preds) %>% arrange(day_of_year)
```

```{r}
df_plot <- data.frame(n = df_test_filtered$number_of_crashes,pred = df_test_filtered$preds, 
                      ymax = df_test_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_test_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])
```

London:

```{r}
i <- 30
df_test_filtered <- df_test %>% filter(police_force == levels(police_force)[i])
preds <- predict(gams[[i]], df_test_filtered, type = "response")
df_test_filtered  <- cbind(df_test_filtered,preds) %>% arrange(day_of_year)
```

```{r}
df_plot <- data.frame(n = df_test_filtered$number_of_crashes,pred = df_test_filtered$preds, 
                      ymax = df_test_filtered$preds + k[i]*S[i,],
                      ymin = pmax( df_test_filtered$preds - k[i]*S[i,], rep(0, 365)),td = 1:365)

df_plot %>% ggplot() + geom_ribbon(aes(x=td,y = pred,ymin = ymin, ymax = ymax), 
                                   fill = "blue", alpha = 0.25) + 
  geom_line(aes(x=td,y = n),linewidth=1,colour = "black") + 
  geom_line(aes(x=td,y = pred),linewidth=1,colour = "blue") + labs(x = "Day", y = "Number of accients",
                                                     title = levels(df_test$police_force)[i])
```



